# Interpreting OLS

```{r}
#| include: false
library(fixest)
N <- 1000; T <- 6
```

An OLS regression returns one number, $\hat\beta$. What that number estimates when effects are heterogeneous, and how each observation contributes to it, depends on the geometry of the sample and on what else is in the model. This chapter derives the implicit weights OLS uses, and follows them across cross-sectional and panel settings.

Two companion notes work through related applications in causal inference: [Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html) and [OLS cannot be ATE](https://xiangao.github.io/blog_book/ols-ate.html).

## OLS as a weighted average

Start with a simple regression of $y$ on $x$ with an intercept:

$$ y_i = \alpha + \beta x_i + e_i $$ {#eq-simple}

The OLS slope is the familiar ratio of a covariance to a variance:

$$
\hat\beta = \frac{\sum_{i=1}^n (x_i - \bar x)(y_i - \bar y)}{\sum_{i=1}^n (x_i - \bar x)^2}
$$ {#eq-slope}

Now give each observation its own slope through the sample centroid $(\bar x, \bar y)$:

$$
s_i \equiv \frac{y_i - \bar y}{x_i - \bar x}
$$ {#eq-si}

Write $d_i \equiv x_i - \bar x$ for the horizontal distance from the centroid, and $w_i \equiv d_i^2$. Multiply and divide each term of @eq-slope by $d_i$ and it rearranges into

$$
\hat\beta = \frac{\sum_{i=1}^n d_i^2 s_i}{\sum_{i=1}^n d_i^2} = \frac{\sum_{i=1}^n w_i s_i}{\sum_{i=1}^n w_i}
$$ {#eq-wavg}

So $\hat\beta$ is an exact weighted average of the deviation slopes $s_i$, and the weight each observation carries is its squared distance from the mean of $x$. Nothing is assumed here; @eq-wavg is algebra.

### What the weights look like

The figures below draw the weights $w_i$ of @eq-wavg as the thickness of each ray from the centroid. They grow with the square of horizontal distance.

```{r}
#| fig-width: 5
#| fig-height: 4
show <- function(x, y, main = "") {
  xb <- mean(x); yb <- mean(y); d <- x - xb; s <- (y - yb) / d; w <- d^2
  plot(x, y, pch = 16, cex = 1.5, xlim = c(-0.5, 11.5), ylim = c(0, 9),
       xlab = "x", ylab = "y", main = main)
  abline(v = xb, lty = 3, col = "grey55")
  segments(xb, yb, x, y, lwd = 0.6 + 9 * w / sum(w), col = "#9E431899")
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")
  points(x, y, pch = 16, cex = 1.5)
  points(xb, yb, pch = 21, bg = "white", col = "#0E6473", lwd = 2.5, cex = 1.7)
  text(x, y, labels = sprintf("%.1f%%", 100 * w / sum(w)),
       pos = ifelse(y >= yb, 3, 1), cex = .85, col = "#9E4318")
}
show(c(1, 3), c(2, 8), "Two Points")
```

With two points the weights are equal, $50\%$ each, and both $s_i$ are just the line through the pair.

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4), c(2, 8, 5), "Three Points")
```

Add a third point at $x=4$ and $\bar x$ moves to $2.67$. The point at $x=3$ now sits almost on top of $\bar x$, so it takes only $2.4\%$ of the weight even though its deviation slope is the largest of the three ($s_2 = 9.0$).

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4, 9), c(2, 8, 5, 6), "Four Points")
```

Add an extreme point at $x=9$ and $\bar x$ moves to $4.25$. That one observation now carries $65\%$ of the total weight, while $x=4$ falls to $0.18\%$ because it has landed near the new centroid. The weights are relative to the whole sample: add data anywhere and the weights everywhere change.

The concentration does not wash out in larger samples.

```{r}
#| fig-width: 6.5
#| fig-height: 5
set.seed(3)
n <- 100
x <- rnorm(n, 5, 2)
y <- 2 * x + rnorm(n, 0, 3)

xb <- mean(x); yb <- mean(y); w <- (x - xb)^2
o <- order(w, decreasing = TRUE)

plot(x, y, pch = 16, cex = 1, col = "grey40", xlab = "x", ylab = "y", main = "N = 100 Points")
abline(v = xb, lty = 3, col = "grey55")
segments(xb, yb, x, y, lwd = 0.3 + 10 * w / max(w), col = "#9E431870")
abline(lm(y ~ x), lwd = 2.5, col = "#0E6473")
points(xb, yb, pch = 21, bg = "white", col = "#0E6473", lwd = 2.5, cex = 1.8)

round(c(`Weight 10 Nearest Mean (%)` = 100 * sum(w[o[91:100]]) / sum(w),
        `Weight 10 Furthest (%)`     = 100 * sum(w[o[1:10]]) / sum(w),
        `OLS Estimate`                = coef(lm(y ~ x))[["x"]]), 3)
```

The $10\%$ of points furthest from $\bar x$ hold $35\%$ of the weight; the $10\%$ closest hold $0.14\%$. A point sitting exactly at $\bar x$ has an undefined $s_i$ and zero weight: the wild value arrives paired with a weight that annihilates it.

## Deviation slopes $s_i$ versus true effects $b_i$

Now let each unit have its own structural slope $b_i$:

$$ y_i = \alpha + b_i x_i + e_i $$ {#eq-het}

$b_i$ is a parameter we never see. $s_i$ from @eq-si is computed from the data. The two are not the same object, and the gap between them is the point of this section.

### Decomposing $s_i$

Let $\bar b = n^{-1}\sum_i b_i$ and $C = n^{-1}\sum_i (b_i - \bar b)(x_i - \bar x)$, the sample covariance between $b$ and $x$. Substituting $x_i = d_i + \bar x$ into @eq-het,

$$
s_i = b_i + \frac{\bar x (b_i - \bar b) - C}{d_i} + \frac{e_i - \bar e}{d_i}
$$ {#eq-decomp}

The middle term is the trouble. When effects are heterogeneous ($b_i \neq \bar b$) it does not vanish, and dividing it by $d_i$ makes $s_i$ blow up for points near $\bar x$ — before any noise enters. Taking expectations over $e_i$,

$$
E[s_i \mid x, b] = b_i + \frac{\bar x (b_i - \bar b) - C}{d_i} \neq b_i
$$ {#eq-esi}

So $s_i$ is biased for $b_i$, observation by observation.

### Aggregating up to $\hat\beta_{OLS}$

Take the $d_i^2$-weighted expectation of $s_i$, combining @eq-wavg and @eq-esi:

$$
E[\hat\beta_{OLS} \mid x, b] = \frac{\sum_i d_i^2 E[s_i \mid x, b]}{\sum_i d_i^2} = \frac{\sum_i d_i^2 b_i + \bar x \sum_i d_i (b_i - \bar b) - C \sum_i d_i}{\sum_i d_i^2}
$$ {#eq-agg}

Because $\sum_i d_i = 0$ and $\sum_i d_i (b_i - \bar b) = n C$, the numerator collapses:

$$
\sum_i d_i^2 E[s_i \mid x, b] = \sum_i d_i^2 b_i + n \bar x C = \sum_i b_i d_i (d_i + \bar x) = \sum_i b_i d_i x_i
$$ {#eq-num}

And since $\sum_i d_i^2 = \sum_i d_i x_i$, we get the exact expectation:

$$
E[\hat\beta_{OLS} \mid x, b] = \frac{\sum_{i=1}^n d_i x_i b_i}{\sum_{i=1}^n d_i x_i} = \frac{\sum_{i=1}^n (x_i - \bar x) x_i b_i}{\sum_{i=1}^n (x_i - \bar x)^2}
$$ {#eq-exact}

The individual biases $E[s_i] - b_i$ of @eq-esi do not cancel. They reshape the weighting instead: from $d_i^2$ on the noisy $s_i$ to $d_i x_i$ on the true slopes $b_i$. Each $s_i$ is wrong in a specific, permanent way, yet combined they are exactly right, because the wrongness moves into the weights rather than vanishing.

### So what?

@eq-exact has three consequences for practice.

1. **OLS is not the ATE under heterogeneity.** $\hat\beta_{OLS}$ is not the plain average effect $\bar b = E[b_i]$. It is a variance-weighted parameter, and it equals the ATE only when the $b_i$ are uncorrelated with the leverage $d_i x_i = (x_i - \bar x)x_i$.
2. **Leverage annihilates the individual bias.** @eq-esi shows $s_i$ is a hopeless estimate of $b_i$ precisely where $d_i \approx 0$. The weight $w_i = d_i^2$ zeroes out exactly those unstable points, turning biased observation-level slopes into the exact parameter of @eq-exact.
3. **The same mechanism runs through the causal estimators.** With a binary regressor it puts the most weight where treatment is near a coin flip (Angrist 1998); grouped by treatment status it reverses the propensity weights, favouring the smaller group (Słoczyński 2022). Both are @eq-exact read at different granularities, as the rest of the chapter shows.

We check the algebra directly.

```{r}
set.seed(6)
n <- 2000
b <- rnorm(n, 1, 0.5)
x <- rnorm(n, 0, 2)
y <- b * x + rnorm(n, 0, 2)
xb <- mean(x); w <- (x - xb)^2
s  <- (y - mean(y)) / (x - xb)

round(cbind(value = c(
  `cor(s_i, b_i)`            = cor(s, b),
  `sd(b_i)`                  = sd(b),
  `sd(s_i)`                  = sd(s),
  `Weighted Avg of s_i`      = sum(w * s) / sum(w),
  `Weighted Avg of b_i`      = sum(w * b) / sum(w),
  `OLS Estimate`             = unname(coef(lm(y ~ x))[2]))), 3)
```

The $s_i$ are wildly dispersed ($\text{sd}(s_i) > 100$) and barely correlated with the true $b_i$, as @eq-esi predicts. Yet their leverage-weighted average matches $\hat\beta_{OLS}$ exactly (@eq-wavg), and the same weighting of the true $b_i$ converges to it too (@eq-exact). The near-zero Pearson correlation is an artefact of the explosive tails — a few points near $\bar x$ dominate it — while the rank correlation on the same data is much higher. The signal is buried, which is not the same as absent.

Noise is not what distorts $s_i$. Turn it off entirely, set $y_i = b_i x_i$, and ask how well each $s_i$ tracks its own $b_i$:

```{r}
y0 <- b * x
s0 <- (y0 - mean(y0)) / (x - xb)
round(c(`cor(s_i, b_i), no noise` = cor(s0, b),
        `Weighted Avg of s_i`     = sum(w * s0) / sum(w),
        `Weighted Avg of b_i`     = sum(w * b) / sum(w),
        `OLS, noiseless`          = unname(coef(lm(y0 ~ x))[2])), 4)
```

The correlation stays at $0.14$ with no noise in the data at all. A ray's slope reflects where every point sits, not just its own, so when the $b_i$ genuinely differ one point's ray cannot recover that point's own slope. Noise makes it worse; it is not the cause.

### Under homogeneity, the weights buy precision

If $b_i = b$ for all $i$, @eq-decomp reduces to $s_i = b + (e_i - \bar e)/d_i$. With homoskedastic errors $\text{Var}(e_i) = \sigma^2$,

$$
E[s_i \mid X] = b, \qquad \text{Var}(s_i \mid X) = \frac{\sigma^2 (1 - 1/n)}{d_i^2} \propto \frac{1}{d_i^2}
$$ {#eq-precision}

Now $w_i = d_i^2$ is exactly the inverse-variance weight. Any weighting of the $s_i$ — equal, $d_i^2$, anything — is unbiased for $b$; what the weights buy is precision, not correctness, and $d_i^2$ is the choice that minimises variance. That is OLS as BLUE. The argument is conditional on $X$, where $d_i$ is a known constant and the weights are functions of $x$ alone.

## The general rule: FWL

The panel chapter met two weightings — by $T_i$ when a panel is unbalanced, and by variance when a regression mixes within and between variation. Neither is a panel quirk. Both follow from one rule.

Regress $y$ on a regressor of interest $V$ and a control matrix $X$, and let $\tilde V$ be the residual from projecting $V$ on $X$. By the Frisch–Waugh–Lovell theorem,

$$
\hat\beta = \frac{\sum_{i=1}^n \tilde V_i y_i}{\sum_{i=1}^n \tilde V_i^2}
$$ {#eq-fwl}

Be careful what $\tilde V_i$ is. A single observation has no variation of its own; variation belongs to a collection. What it has is a *deviation* — how far its regressor sits from what the rest of the model predicts for it. With only an intercept, that is distance from the grand mean. With unit fixed effects, it is distance from the unit's own mean. Which mean gets subtracted is the most consequential detail in this section, and we come back to it.

@eq-fwl reads two ways.

1. **Weights on outcomes.**
   $$ \hat\beta = \sum_{i=1}^n \omega_i y_i, \qquad \omega_i = \frac{\tilde V_i}{\sum_{j=1}^n \tilde V_j^2} $$ {#eq-omega}
   The $\omega_i$ are signed, sum to zero, and say how each observed $y_i$ moves $\hat\beta$.

2. **Weights on effects.** Under a heterogeneous model $y_i = \beta_i V_i + X_i \gamma + e_i$,
   $$ \hat\beta = \frac{\sum_{i=1}^n w_i \beta_i}{\sum_{i=1}^n w_i}, \qquad w_i = \tilde V_i V_i $$ {#eq-weff}
   When $X$ holds saturated indicators or fixed effects, $\sum_i \tilde V_i V_i = \sum_i \tilde V_i^2 \ge 0$.

| | Weights on outcomes $\omega_i$ | Weights on effects $w_i$ |
|---|---|---|
| Weights what | the observed $y_i$ | that observation's own effect $\beta_i$ |
| Sign | signed, sums to zero | non-negative |
| Computable | yes, $y$ is observed | no, the $\beta_i$ are never seen |
| Holds when | always, it is algebra | after writing effects into $y$; the per-unit reading also needs saturation |
| Answers | which rows drive this number, and does any enter with the wrong sign | is this the effect I meant, and whose effect is it mostly |

These are different objects. In @eq-wavg the weight $d_i^2$ sat on $s_i$, because $s_i$ was what we were averaging; here $w_i = \tilde V_i V_i$ sits on $\beta_i$. They differ observation by observation and coincide only when $x$ is centred, though they always sum to the same total, since $\sum_i d_i x_i = \sum_i d_i^2$. Default to the effect reading: nearly every question that sends people here — why do smaller groups get more weight — is a question about the estimand. Reach for the outcome reading when overlap is in doubt and rows may enter with the wrong sign, which squares cannot reveal.

### Where the square comes from

The square in $w_i$ is derived, not assumed. Substitute $y_i = \beta_i V_i + X_i \gamma + e_i$ into the numerator of @eq-fwl and it becomes $\sum_i \beta_i \tilde V_i V_i$; over the whole sample,

$$
\sum_{i=1}^n \tilde V_i V_i = \sum_{i=1}^n \tilde V_i^2
$$ {#eq-square}

because $\tilde V$ is orthogonal to the fitted part of $V$.

In a panel, several rows share a unit, and it is natural to ask whether the same cancellation holds *within a single unit* — summed over that unit's own $t$'s — and not only over the whole sample. Write $V_{it}$ for the observation and $\bar V_i$ for the unit's own mean. Demeaning gives the within-unit version by direct expansion,

$$
\sum_{t=1}^{T} (V_{it} - \bar V_i) V_{it} = \sum_{t=1}^{T} V_{it}^2 - T \bar V_i^{\,2} = \sum_{t=1}^{T} (V_{it} - \bar V_i)^2
$$ {#eq-within-square}

while a coarser control matches only in total, not unit by unit. Hazlett and Shinkre (2024) make this within-unit cancellation the centre of their treatment.

### Special cases of the effect weights

In general $\sum_t \tilde V_{it}^2$ within a unit does not factor into anything tidier — $\tilde V_{it}$ carries both subscripts even when $V$ does not. What changes from model to model is only whether it happens to simplify:

| Model | Regressor $V$ | Within-unit weight $w_i$ |
|---|---|---|
| Panel, group-level regressor | $z_i$ (unit-level) | $T_i \tilde z_i^2$ |
| Panel, within-unit variation | $x_{it}$ (time-varying) | $(T-1)\, \widehat{\text{Var}}_i(x)$ |
| Binary treatment, saturated controls | $D_i \in \{0, 1\}$ | $n_x \hat D(X)(1 - \hat D(X))$ |

Add a time-varying $x_{it}$ alongside $z_i$ in the first row and the shortcut is gone: residualising $z_i$ on something that moves within the unit makes $\tilde z$ move within the unit too. The rule survives, the shortcut does not, and the $T_i$ form is the one-regressor special case. The last row is Angrist (1998), whose strata weights $\hat D(x)(1 - \hat D(x))$ peak where treatment is closest to a coin flip. Same rule, different regressor.

## Two causal decompositions: Angrist and Słoczyński

Both results are usually met as free-standing facts about treatment effects. They are the rule above, read at two granularities. @eq-weff says the coefficient is a weighted average of effects with weights $w_i = \tilde V_i V_i$; everything below only evaluates those weights. Angrist reads them at the stratum level. Słoczyński reads the same weights one level up, grouped by treatment status, where they come out reversed relative to the natural shares of the ATE. The two subsections are the same calculation, twice.

Let $D \in \{0, 1\}$ be a binary treatment and $X$ discrete strata $x \in \{1, \dots, K\}$.

### Angrist (1998): weighting by stratum

Residualising $D_i$ on a full set of stratum indicators gives $\tilde D_i = D_i - \hat D(x)$, where $\hat D(x) = E[D \mid X = x]$. Sum the effect weights $w_i = \tilde D_i D_i$ within a stratum:

$$
w_x = \sum_{i \in x} \tilde D_i D_i = \sum_{i \in x} D_i^2 - \hat D(x) \sum_{i \in x} D_i = n_x \hat D(x) - n_x \hat D(x)^2 = n_x \hat D(x)(1 - \hat D(x))
$$ {#eq-angrist-w}

Nothing new has been added: it is $w_i = \tilde V_i V_i$ from @eq-weff with a binary $V$, where $D_i^2 = D_i$ makes the sum collapse. The maximum at $\hat D(x) = 1/2$ is just the fact that a binary variable has most variance when it is a coin flip.

Note what plays the role of $s_i$ here. A single observation with a binary regressor has no slope; you need one treated and one control unit to form a difference. The smallest unit carrying an effect is therefore the *stratum*, and $s_i$ becomes the stratum difference in means. The regressor sets the granularity, we do not choose it.

Let $\hat\delta_x$ be the stratum difference-in-means. Substituting @eq-angrist-w into @eq-weff,

$$
\hat\beta_{OLS} = \frac{\sum_{x=1}^K n_x \hat D(x)(1 - \hat D(x))\, \hat\delta_x}{\sum_{x=1}^K n_x \hat D(x)(1 - \hat D(x))}
$$ {#eq-angrist}

so strata with propensity near $\hat D(x) = 0.5$ get the most weight per observation.

```{r}
set.seed(21)
n <- 400000
X <- sample(1:5, n, TRUE)

compare <- function(px) {
  d  <- rbinom(n, 1, px[X]); ta <- c(0, 1, 2, 3, 4)[X]
  yy <- ta * d + X + rnorm(n)
  dt <- resid(lm(d ~ factor(X)))
  nx <- as.vector(table(X)); dh <- as.vector(tapply(d, X, mean))
  DIM <- as.vector(tapply(yy[d == 1], X[d == 1], mean) - tapply(yy[d == 0], X[d == 0], mean))
  rho <- mean(d); ATT <- weighted.mean(ta, d); ATU <- weighted.mean(ta, 1 - d)
  c(`rho`                      = rho,
    `Max |Weight Diff|`         = max(abs(as.vector(tapply(dt * d, X, sum)) - nx*dh*(1-dh))),
    `OLS Saturated`            = unname(coef(lm(yy ~ d + factor(X)))[["d"]]),
    `Angrist Weighted`         = sum(nx*dh*(1-dh) * DIM) / sum(nx*dh*(1-dh)),
    `Sloczynski Approx`        = (1 - rho) * ATT + rho * ATU,
    `ATE`                      = mean(ta))
}
res <- cbind(`Balanced (rho ~ 0.5)`   = compare(c(.08, .20, .50, .75, .93)),
             `Unbalanced (rho ~ 0.09)` = compare(c(.02, .04, .08, .12, .20)))
round(res[-2, ], 4)
```

The two weight vectors — the stratum sums $\sum \tilde D_i D_i$ and Angrist's $n_x \hat D(x)(1 - \hat D(x))$ — agree to within $6 \times 10^{-10}$ in both designs, and the Angrist-weighted average reproduces the OLS coefficient exactly. In the balanced design the coefficient is $2.012$ against an ATE of $1.998$: close, because balance is where the weighting does least. Push the treated share to $0.09$ and the coefficient jumps to $2.874$ against the same $1.998$. Nothing about the effects changed. Only the treated share did.

### Słoczyński (2022): aggregating to two groups

Angrist sorts by covariate stratum. Słoczyński sorts by treatment status instead, collapsing everything into two numbers — the effect on the treated and the effect on the untreated. Aggregating the stratum effects $\hat\delta_x$ of @eq-angrist with each stratum's count of treated or untreated units gives

$$
\text{ATT} = \frac{\sum_x n_x \hat D(x)\, \hat\delta_x}{\sum_x n_x \hat D(x)}, \qquad
\text{ATU} = \frac{\sum_x n_x (1 - \hat D(x))\, \hat\delta_x}{\sum_x n_x (1 - \hat D(x))}
$$ {#eq-attatu}

and the ATE combines them with their natural shares, $\rho = P(D = 1)$:

$$
\text{ATE} = \rho\, \text{ATT} + (1 - \rho)\, \text{ATU}
$$ {#eq-ate}

The OLS weights of @eq-angrist go as the propensity *variance* $\hat D(x)(1 - \hat D(x))$; the group weights of @eq-attatu go as the propensity itself or its complement. The relation between the two is Słoczyński's result. Write $p = \hat D(X)$ for the cell propensity. Because the design is saturated, residualising $D$ on the strata leaves exactly $\tilde D = D - p$, so the coefficient is $E[\tilde D\, Y]/E[p(1-p)]$. Since $p$ is constant within a stratum, $Y$ enters only through the group means $m_j(p) = E[Y \mid p, D = j]$, and, using $\tilde D = 1 - p$ for treated and $-p$ for untreated units,

$$
\beta_{OLS} = \frac{\rho\, E[(1 - p)\, m_1(p) \mid D = 1] - (1 - \rho)\, E[p\, m_0(p) \mid D = 0]}{E[p(1 - p)]}
$$ {#eq-sloc1}

Substitute the within-group linear projections $m_j(p) = \alpha_j + \gamma_j p + \varepsilon_j(p)$. The residual term $E[p(1-p)(\varepsilon_1 - \varepsilon_0)]/E[p(1-p)]$ vanishes when the group means are linear in $p$, leaving

$$
\beta_{OLS} = (\alpha_1 - \alpha_0) + (\gamma_1 - \gamma_0)\, \frac{E[p^2(1 - p)]}{E[p(1 - p)]}
$$ {#eq-sloc2}

The last ratio is moment algebra. With $\mu_j = E[p \mid D = j]$ and $V_j = \text{Var}(p \mid D = j)$,

$$
\frac{E[p^2(1 - p)]}{E[p(1 - p)]} = w_1 \mu_1 + w_0 \mu_0
$$ {#eq-moment}

holds identically in the distribution of $p$. Under linearity $\text{ATT} = (\alpha_1 - \alpha_0) + (\gamma_1 - \gamma_0)\mu_1$ and $\text{ATU} = (\alpha_1 - \alpha_0) + (\gamma_1 - \gamma_0)\mu_0$, so @eq-sloc1 through @eq-moment combine into the main result,

$$
\hat\beta_{OLS} = w_1\, \text{ATT} + w_0\, \text{ATU}, \qquad
w_1 = \frac{(1 - \rho) V_0}{\rho V_1 + (1 - \rho) V_0}, \quad w_0 = 1 - w_1
$$ {#eq-sloc-main}

where $V_j$ is the variance of the propensity score within group $j$. This is exact for the effects of the within-group linear projections; for ATT and ATU it is exact when the group means are linear in $p$ and an approximation otherwise.

Two things about @eq-sloc-main matter. First, the weight $w_1$ on ATT falls as the treated share $\rho$ rises: the larger the treated group, the smaller its say. That is the whole of "smaller groups get more weight." Second, when the propensity has equal variance in the two groups, $V_1 = V_0$, the weights collapse to the reversed shares,

$$
\hat\beta_{OLS} \approx (1 - \rho)\, \text{ATT} + \rho\, \text{ATU}
$$ {#eq-reversed}

the same two quantities as the ATE in @eq-ate, with the weights **swapped**. At $\rho = 0.1$ the coefficient puts $0.9$ on the effect for the treated tenth; the ATE puts $0.1$ there.

One caveat. @eq-reversed is exact under equal propensity variance across groups — not "no covariates," though no covariates is one trivial way to get there. Słoczyński calls the condition unlikely to hold exactly and offers $\rho$ as a rule of thumb, cleanest near $\rho = 0\%$, $50\%$, or $100\%$. In the simulation above the raw-$\rho$ form gives $2.029$ against an exact $2.012$, and $2.858$ against an exact $2.874$: close enough to reason with, not close enough to quote.

### The three, side by side

| Decomposition | Sorted by | Smallest effect unit | Exact when | Reference |
|---|---|---|---|---|
| Observation rule | nothing | $s_i$ | always (identity) | @eq-wavg |
| Angrist (1998) | covariate stratum $x$ | stratum difference $\hat\delta_x$ | saturated controls | @eq-angrist |
| Słoczyński (2022) | treatment status $D$ | group averages ATT, ATU | equal propensity variance | @eq-reversed |

Read down the last column. The observation identity always holds and says little. Angrist's needs saturation and buys an interpretable weight. Słoczyński's needs more and buys the sharpest statement of the three — that the coefficient and the ATE are the same two numbers with the weights reversed. Underneath all of them, Hazlett and Shinkre (2024) show these weights are exactly the bias from fitting a constant effect when effects vary, and argue for estimators that tolerate heterogeneity rather than for diagnosing the weights after the fact.

## Within versus between, in panels

Pooled OLS mixes two sources of variation: within-unit, the deviations from a unit's own mean, and between-unit, the differences across unit means. Which one dominates is not chosen; it follows from the same rule. Take the within side first.

### Within-unit weighting: an extreme case

Half the units barely move ($x$ has sd $0.2$ around their own mean) and have a true slope of zero. The other half move a lot (sd $3$) and have a true slope of two. The average unit's slope is one.

```{r}
set.seed(1)
half <- N / 2
g <- data.frame(id   = rep(1:N, each = T),
                move = rep(c(rep(0.2, half), rep(3, half)), each = T),
                b    = rep(c(rep(0,   half), rep(2, half)), each = T))
g$x <- rnorm(N * T, 0, g$move)
g$y <- g$b * g$x + rep(rnorm(N), each = T) + rnorm(N * T)

wg <- tapply(resid(feols(x ~ 1 | id, g))^2, g$id, sum)
round(cbind(value = c(
  `average of the two true slopes`      = mean(c(0, 2)),
  `FE estimate`                         = coef(feols(y ~ x | id, g))[["x"]],
  `weight held by the steady half (%)`  = 100 * sum(wg[1:half]) / sum(wg))), 3)
```

The fixed-effects estimate is $1.98$, because the five hundred steady units hold under half of one percent of the weight. Their six observations sit almost on top of their own unit mean — with unit effects that is the relevant deviation — and squaring puts them nowhere in the sum. Nothing has gone wrong: the zero-slope units were never really in the regression.

### Splitting the two channels

That result was about the *within-unit* deviation. Plain pooled OLS uses distance from the grand mean instead, and the two coincided above only because both halves were built around zero. Separate them and the algebra says exactly what happens.

Take two equal-sized groups $G \in \{1, 2\}$ with means $\mu_g = E[x \mid G=g]$ and $\bar y_g = E[y \mid G=g]$, within-group variances $\sigma_g^2$, and true slopes $\beta_g$. By the law of total covariance,

$$
\text{Cov}(x, y) = E[\text{Cov}(x, y \mid G)] + \text{Cov}(E[x \mid G], E[y \mid G])
$$ {#eq-totcov}

$$
E[\text{Cov}(x, y \mid G)] = \tfrac{1}{2}(\beta_1 \sigma_1^2 + \beta_2 \sigma_2^2)
$$ {#eq-within-cov}

$$
\text{Cov}(E[x \mid G], E[y \mid G]) = \tfrac{1}{4}(\mu_1 - \mu_2)(\bar y_1 - \bar y_2) = \tfrac{1}{4}\Delta\mu\, \Delta\bar y
$$ {#eq-between-cov}

and by the law of total variance,

$$
\text{Var}(x) = \tfrac{1}{2}(\sigma_1^2 + \sigma_2^2) + \tfrac{1}{4}\Delta\mu^2
$$ {#eq-totvar}

Dividing @eq-totcov by @eq-totvar gives the asymptotic pooled slope as a convex combination of a within and a between slope,

$$
\text{plim}\;\hat\beta_{OLS} = \frac{\tfrac{1}{2}(\beta_1 \sigma_1^2 + \beta_2 \sigma_2^2) + \tfrac{1}{4}\Delta\mu\,\Delta\bar y}{\tfrac{1}{2}(\sigma_1^2 + \sigma_2^2) + \tfrac{1}{4}\Delta\mu^2} = \frac{V_{within}\, \beta_{within} + V_{between}\, \beta_{between}}{V_{within} + V_{between}}
$$ {#eq-pooled}

with

$$
\beta_{within} = \frac{\beta_1 \sigma_1^2 + \beta_2 \sigma_2^2}{\sigma_1^2 + \sigma_2^2}, \qquad V_{within} = \tfrac{1}{2}(\sigma_1^2 + \sigma_2^2)
$$ {#eq-within-comp}

$$
\beta_{between} = \frac{\Delta\bar y}{\Delta\mu}, \qquad V_{between} = \tfrac{1}{4}\Delta\mu^2
$$ {#eq-between-comp}

Two features are worth pausing on. $\beta_{within}$ is itself a variance-weighted average of the two slopes — the same $\sigma^2$ weighting as everywhere else in this chapter, one level up. And $\beta_{between}$ is a two-point slope, rise over run between the group centres, so it can take any value at all regardless of what happens inside either group. The $1/4$ is the only piece specific to equal group sizes; with shares $p$ and $1-p$ it becomes $p(1-p)$, Angrist's weight again.

Fixed effects removes $V_{between}$ by demeaning and isolates $\beta_{within}$. As the groups pull apart, $\Delta\mu \to \infty$, $V_{between}$ swamps $V_{within}$ and drags $\hat\beta_{OLS}$ toward $\beta_{between}$.

```{r}
b_within <- (0 * 0.2^2 + 2 * 3^2) / (0.2^2 + 3^2)
V_within <- 0.5 * (0.2^2 + 3^2)
predicted <- function(shift) V_within * b_within / (V_within + 0.25 * shift^2)

park <- function(shift) {
  set.seed(1)
  half <- N / 2
  move <- c(rep(0.2, half), rep(3, half))
  b    <- c(rep(0,   half), rep(2, half))
  x    <- rnorm(N * T, 0, rep(move, each = T)) + rep(c(rep(shift, half), rep(0, half)), each = T)
  y    <- rep(b, each = T) * x + rep(rnorm(N), each = T) + rnorm(N * T)
  d    <- data.frame(id = rep(1:N, each = T), x = x, y = y)
  c(`FE (theory 1.991)` = coef(feols(y ~ x | id, d))[["x"]],
    `OLS Predicted`     = predicted(shift),
    `OLS Actual`        = coef(lm(y ~ x, d))[["x"]])
}

round(t(sapply(c(`shift 0` = 0, `shift 10` = 10, `shift 100` = 100), park)), 3)
```

Fixed effects stays flat near the theoretical $\beta_{within} = 1.991$ wherever the steady group sits. Plain OLS falls from $1.983$ to $0.004$ on identical true slopes: as $\Delta\mu$ grows, the between weight $\tfrac{1}{4}\Delta\mu^2$ swamps the fixed within weight $4.52$, and the coefficient is dragged toward $\beta_{between} = 0$ like $9/(4.52 + \Delta\mu^2/4)$, the numerator $\tfrac{1}{2}(\beta_1 \sigma_1^2 + \beta_2 \sigma_2^2) = 9$.

```{r}
#| fig-width: 9
#| fig-height: 4.2
set.seed(5)
m <- 200
draw <- function(shift, main) {
  xw <- rnorm(m, 0, 3);       yw <- 2 * xw + rnorm(m)
  xs <- rnorm(m, shift, 0.2); ys <- rnorm(m)
  x <- c(xw, xs); y <- c(yw, ys)
  plot(x, y, pch = 16, cex = .5, col = c(rep("#9E431855", m), rep("#0E6473", m)),
       xlab = "x", ylab = "y", main = main)
  abline(v = mean(x), lty = 3, col = "grey50")
  segments(mean(xw), mean(yw), mean(xs), mean(ys), lty = 2, lwd = 2, col = "grey30")
  points(c(mean(xw), mean(xs)), c(mean(yw), mean(ys)), pch = 21, bg = "white", cex = 1.4)
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")
  dmu <- mean(xs) - mean(xw)
  Vb  <- 0.25 * dmu^2
  Vw  <- 0.5 * (var(xs) + var(xw))
  c(`Within Slope 1`     = unname(coef(lm(yw ~ xw))[2]),
    `Within Slope 2`     = unname(coef(lm(ys ~ xs))[2]),
    `Between Slope`      = unname((mean(ys) - mean(yw)) / dmu),
    `Between Weight (%)` = 100 * Vb / (Vb + Vw),
    `OLS Estimate`       = unname(coef(lm(y ~ x))[2]))
}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
a <- draw(0,  "Centroids Coincide (Delta mu = 0)")
b <- draw(10, "Separated Centroids (Delta mu = 10)")
par(op)
round(cbind(`Delta mu = 0` = a, `Delta mu = 10` = b), 3)
```

Read the two columns. The slopes inside each group barely change — about $2$ for the wobbly cloud, indistinguishable from zero for the steady strip (the $-0.288$ is noisy, since a regression on an $x$ that hardly varies has a standard error near $0.35$). What changes is the weight on the between comparison: $0.03\%$ when the centres coincide, $85\%$ when they are ten apart. So the fit lands on $2.01$ in the first panel and $0.29$ in the second.

Two details earn a name. The between slope reported on the left, $0.882$, is meaningless — and harmlessly so: with the centres nearly on top of each other $\Delta\mu \approx 0$, so an unstable ratio is multiplied by a vanishing weight and contributes nothing. And the strip has influence on the right despite barely moving: influence is distance from the mean of $x$, and the strip sits ten units from it. What it contributes is not its slope but its position — it anchors one end of the dashed line. A group dummy removes that channel and returns the within answer.

So "the weight is how much $x$ moves" is the demeaned case only. It is exact once unit effects are removed, and false in general, because with the halves at different levels of $x$ part of the fitted line is a comparison *between* them rather than a weighting of within-unit slopes.

### When the weighting changes nothing

Weighting only distorts when the effect is related to the weight. Give every unit its own slope, let units differ in how much $x$ moves, and compare two designs: in (A) the slopes are unrelated to that movement, in (B) the slope is larger where $x$ moves more. Both average near one. Design (A) is the $s_i$-versus-$b_i$ story from before, run as a panel.

```{r}
set.seed(7)
sdX <- runif(N, 0.2, 3)

panel <- function(b) {
  h   <- data.frame(id = rep(1:N, each = T),
                    x  = rnorm(N * T, 0, rep(sdX, each = T)))
  h$y <- rep(b, each = T) * h$x + rep(rnorm(N), each = T) + rnorm(N * T)
  h
}
compareAB <- function(h, b) {
  w <- tapply(resid(feols(x ~ 1 | id, h))^2, h$id, sum)
  c(`FE estimate`            = coef(feols(y ~ x | id, h))[["x"]],
    `true slopes, plain avg` = mean(b),
    `true slopes, weighted`  = sum(w * b) / sum(w))
}

b1 <- rnorm(N, 1, 1)
b2 <- 1 + 0.8 * (sdX - 1.6)
h1 <- panel(b1); h2 <- panel(b2)

round(cbind(`(A) unrelated` = compareAB(h1, b1),
            `(B) aligned`   = compareAB(h2, b2)), 4)
```

In (A) all three numbers agree to within sampling noise: the weighting is there and costs nothing. In (B) the estimate is $1.53$ while the plain average of the true slopes is $1.01$, and the weighted average, $1.54$, is what the estimator was tracking all along. None of this is a defect — a unit whose $x$ barely moves genuinely says little about how $y$ responds, and leaning on the units that do move is efficient. Design (B) is the continuous version of what Słoczyński describes for a binary treatment.

## Summary

1. **Leverage weighting.** OLS weights each observation by regressor variance $\tilde V_i^2 = (x_i - \bar x)^2$ (@eq-wavg). Units that vary more dominate the slope.
2. **What is being estimated.** Under heterogeneous slopes $b_i$, OLS estimates the variance-weighted parameter $\sum_i w_i b_i / \sum_i w_i$ (@eq-exact). It equals the ATE only when the $b_i$ are uncorrelated with the weights.
3. **Between bias in pooled OLS.** Without fixed effects, group-mean differences $\Delta\mu$ enter through $V_{between}\,\beta_{between}$ (@eq-pooled) and distort the within-unit parameter; demeaning removes them.
4. **When it matters.** The weighting always happens. It changes the answer only when effects are heterogeneous *and* the heterogeneity aligns with the weight. With homogeneous effects it is harmless. With heterogeneous effects the question is never whether OLS is weighting — only whether we know what it weighted.
