# Interpreting OLS

```{r}
#| include: false
library(fixest)
N <- 1000; T <- 6
```

A regression returns one number.  What that number *is* --- whose effect
it reflects, which observations produced it, whether it is the average
anyone had in mind --- is a separate question from how it is computed,
and the answer is less obvious than it looks.

This chapter collects the material.  It sits at the end because it
draws on the between-versus-within split from
[Panel Data Models](panel-data.qmd), but almost nothing in it is
specific to panels: the same rule governs a plain cross-sectional
regression, and the binary-treatment case that most of the applied
literature discusses is one line of the table below.

Two companion pieces develop the same ideas with a treatment-effects
emphasis:
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html)
and
[OLS cannot be ATE](https://xiangao.github.io/blog_book/ols-ate.html).

## Start with two points

Begin as small as possible.  Two points, and a regression line through
them.  Everyone knows what that line is: it joins them.

Now describe it in a way that will keep working when there are more
points.  Take the **centroid** $(\bar x, \bar y)$ and, for each point,
the slope of the line running from the centroid out to it:

$$
s_i \;=\; \frac{y_i - \bar y}{x_i - \bar x} .
$$

Call $s_i$ that observation's own slope.  With two points the centroid
lies on the segment joining them, so both rays *are* that segment and
both $s_i$ are the same number.

Attach a weight to each, $d_i^2=(x_i-\bar x)^2$, the squared horizontal
distance from the centroid.  We will justify that choice shortly; for
now just take the weighted average.

In the figures below the hollow circle is the centroid, the dotted
vertical line is $\bar x$, the rust rays run from the centroid to each
point with **thickness proportional to that point's weight**, and the
solid line is the fit.  Percentages label each point's share of the
total weight.

```{r}
#| fig-width: 5
#| fig-height: 4
show <- function(x, y, main = "") {
  xb <- mean(x); yb <- mean(y); d <- x - xb; s <- (y - yb) / d; w <- d^2
  plot(x, y, pch = 16, cex = 1.5, xlim = c(-0.5, 11.5), ylim = c(0, 9),
       xlab = "x", ylab = "y", main = main)
  abline(v = xb, lty = 3, col = "grey55")                      # the mean of x
  segments(xb, yb, x, y, lwd = 0.6 + 9 * w / sum(w),           # rays, thickness = weight
           col = "#9E431899")
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")                  # the fit
  points(x, y, pch = 16, cex = 1.5)
  points(xb, yb, pch = 21, bg = "white", col = "#0E6473", lwd = 2.5, cex = 1.7)
  text(x, y, labels = sprintf("%.1f%%", 100 * w / sum(w)),   # label away from centroid
       pos = ifelse(y >= yb, 3, 1), cex = .85, col = "#9E4318")
  print(round(data.frame(x, y, s_i = s, weight = w,
                         `weight %` = 100 * w / sum(w), check.names = FALSE), 3),
        row.names = FALSE)
  cat(sprintf("  centroid x = %.3f | weighted avg of s_i = %.4f | OLS = %.4f\n",
              xb, sum(w * s) / sum(w), coef(lm(y ~ x))[["x"]]))
}
show(c(1, 3), c(2, 8), "two points")
```

Both $s_i$ equal 3, the weights split evenly, and the average is 3 ---
which is the fitted slope.  Note what this means: with two points the
weights are *irrelevant*.  Any weighting of two identical numbers
returns that number.

## Add a third point

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4), c(2, 8, 5), "three points")
```

Now the three own-slopes disagree: 1.8, 9.0 and 0.0.  The fit has to
choose among them, and it does so by weight --- and the weights are
wildly uneven.  The point at $x=3$ has the steepest own-slope of the
three, 9.0, and receives 2.4% of the say, because $x=3$ sits almost on
the centroid at 2.667.  The answer, 1.2857, is the weighted average.

This is the whole mechanism.  A point close to $\bar x$ has almost no
horizontal leverage, so however dramatic its own slope, it barely moves
the line.

## Add a fourth, far out

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4, 9), c(2, 8, 5, 6), "four points")
```

Two things happened at once, and both are worth naming.

The new point at $x=9$ took **65% of the weight by itself**, because
weight is squared distance and it is far from centre.  And the point at
$x=4$, which had 38% of the say a moment ago, collapsed to **0.18%** ---
not because it changed, but because the centroid moved out to 4.25 and
landed almost on top of it.

So weight is not a property of an observation.  It is a property of an
observation *relative to the rest of the sample*, and adding data
anywhere rewrites it everywhere.  The fitted slope fell from 1.29 to
0.31 without a single existing point being touched.

## The identity behind it

None of that was a convention.  Since
$(x_i-\bar x)(y_i-\bar y)=(x_i-\bar x)^2 s_i$, the ordinary OLS formula
*is* that weighted average:

$$
\hat\beta
=\frac{\sum_i (x_i-\bar x)(y_i-\bar y)}{\sum_i (x_i-\bar x)^2}
=\frac{\sum_i (x_i-\bar x)^2\, s_i}{\sum_i (x_i-\bar x)^2}.
$$

Every observation carries a slope of its own, and the fit averages them
with weights $d_i^2$.  The two-group example later in this chapter is
this same arithmetic with three thousand points on each side instead of
two.

One consequence is worth stating before we go on.  An observation
sitting exactly at $\bar x$ has an undefined $s_i$ and zero weight ---
the wild value arrives paired with a weight that annihilates it.  The
four-point panel shows this happening in miniature.

## Do not mistake $s_i$ for $b_i$

This is where care is needed.  $s_i$ is *observable* --- it is computed
from the data.  The structural $b_i$ is not.  It is tempting to treat
the first as a measurement of the second, and at the level of a single
observation that is badly wrong.

```{r}
set.seed(6)
n <- 2000
b <- rnorm(n, 1, 0.5)                 # TRUE slope of each observation, unobservable
x <- rnorm(n, 0, 2)
y <- b * x + rnorm(n, 0, 2)
xb <- mean(x); w <- (x - xb)^2
s  <- (y - mean(y)) / (x - xb)        # OBSERVABLE slope to the centroid

round(cbind(value = c(
  `correlation of s_i with b_i` = cor(s, b),
  `sd of true b_i`              = sd(b),
  `sd of observed s_i`          = sd(s),
  `weighted average of s_i`     = sum(w * s) / sum(w),
  `weighted average of b_i`     = sum(w * b) / sum(w),
  `OLS estimate`                = unname(coef(lm(y ~ x))[2]))), 3)
```

The Pearson correlation is 0.019 and the spread of the observed $s_i$ is
more than two hundred times the spread of the true $b_i$, because $s_i$
divides a noisy vertical gap by a horizontal distance that is sometimes
almost zero.  An observation with an ordinary $b_i$ near one can easily
show an $s_i$ of eighteen.

Do not over-read that 0.019, though.  It is a Pearson correlation on a
variable with enormous tails, so a handful of exploded $s_i$ dominate
it; the rank correlation on the same data is far higher.  The signal is
present, and buried, which is a different thing from absent.

Yet the weighted averages agree: $0.950$ for the $s_i$, $0.992$ for the
$b_i$, and the OLS estimate is $0.950$ --- identical to the $s_i$
version by construction, and close to the $b_i$ version because the
noise averages out.


### Why using $s_i$ works anyway

This ought to look like a problem.  We want the $b_i$; we are averaging
something that barely correlates with them.  The resolution is that the
weight is precisely the right correction.

Everything here is **conditional on $X$**, as usual in regression, and
that matters twice over: it makes $d_i=x_i-\bar x$ a known constant
rather than a random variable, and the weights are functions of $x$
alone, so the whole weighting story only lives in the conditional world.

With homogeneous truth $b$, $y_i-\bar y = b\,d_i + (e_i-\bar e)$, so
$s_i = b + (e_i-\bar e)/d_i$ and $s_i$ is unbiased.  Since
$\text{Var}(e_i-\bar e)=\sigma^2-2\sigma^2/n+\sigma^2/n=\sigma^2(1-1/n)$,

$$
\text{Var}(s_i)\;=\;\frac{\sigma^2\,(1-1/n)}{d_i^{\,2}}
\;\approx\;\frac{\sigma^2}{d_i^{\,2}},
\qquad\text{so}\qquad
d_i^{\,2} \;\propto\; \frac{1}{\text{Var}(s_i)} .
$$

The same $\bar e$ also makes the $s_i$ *dependent*.  From
$\text{Cov}(e_i-\bar e,\,e_j-\bar e)=-\sigma^2/n$,

$$
\text{Cov}(s_i,s_j)\;=\;-\,\frac{\sigma^2}{n\,d_i d_j}\,,
\qquad i\neq j .
$$

Small, of order $1/n$, but not zero.  So $d_i^2$ is the inverse variance
of $s_i$, while the $s_i$ themselves are mildly correlated --- which is
why the efficiency of OLS rests on Gauss--Markov and not on any argument
about combining independent estimates.

The weight is the inverse variance of the thing it weights.  Check it by
splitting on $|d_i|$: the product $d_i^2\,\text{Var}(s_i)$ should be
constant at $\sigma^2$.

```{r}
set.seed(12)
nn <- 200000
xx <- rnorm(nn, 0, 2); yy <- 1 * xx + rnorm(nn, 0, 2)   # true b = 1, sigma^2 = 4
dd <- xx - mean(xx); ss <- (yy - mean(yy)) / dd; ww <- dd^2

bin <- cut(abs(dd), quantile(abs(dd), 0:6/6), include.lowest = TRUE)
print(round(t(sapply(split(seq_len(nn), bin), function(i)
  c(`mean |d|` = mean(abs(dd[i])), `var(s)` = var(ss[i]),
    `mean d^2` = mean(ww[i]), `d^2 * var(s)` = mean(ww[i]) * var(ss[i])))), 2))

round(c(`unweighted mean of s_i` = mean(ss),
        `weighted mean of s_i`   = sum(ww * ss) / sum(ww)), 3)
```

Ignore the first row for a moment.  Across the rest the last column sits
at about 4, which is $\sigma^2$ --- so OLS is the inverse-variance
weighted average of $n$ unbiased estimates of the slope.  The $s_i$ that look absurd are absurd *because* their
$x$ lies near the mean, and that is exactly the quantity the weight
divides out.  The cancellation is algebraic, not approximate:
$d_i^2\, s_i = d_i^2 \cdot (y_i-\bar y)/d_i = d_i(y_i-\bar y)$.

The first row is the exception and it is instructive.  There $|d_i|$
reaches $10^{-10}$, so $s_i$ is a ratio with a vanishing denominator and
has no finite variance at all.  Those observations get weight near zero
and do no damage --- but it means the weighting is not merely efficient.
Without it the estimator would not exist: the *unweighted* mean of the
$s_i$ in this sample is $-23{,}816$, against $0.999$ for the weighted
one.

### How much of this is the noise?

Naturally one asks whether the disconnect is just a large error
variance.  Largely yes.  Shrink $\sigma_e$ and $s_i$ closes on $b_i$.

```{r}
set.seed(9)
nn <- 20000
bb <- rnorm(nn, 1, 0.5); xx <- rnorm(nn, 0, 2)
t(sapply(c(2, 1, 0.5, 0.1, 0.01, 0), function(se) {
  yy  <- bb * xx + rnorm(nn, 0, se)
  dd  <- xx - mean(xx); ss <- (yy - mean(yy)) / dd
  far <- abs(dd) > quantile(abs(dd), .5)          # drop the half nearest the mean
  round(c(sigma_e = se, pearson = cor(ss, bb),
          spearman = cor(ss, bb, method = "spearman"),
          `pearson, far half` = cor(ss[far], bb[far])), 3)
}))

# with no noise at all, how far is s_i from b_i?
dd <- xx - mean(xx); ss <- (bb * xx - mean(bb * xx)) / dd
far <- abs(dd) > quantile(abs(dd), .5)
round(c(`max |s-b|, all`      = max(abs(ss - bb)),
        `max |s-b|, far half` = max(abs((ss - bb)[far]))), 4)
```

The rank correlation climbs from 0.25 to 0.997, and among observations
away from the mean the Pearson correlation reaches 1.000.  So yes: with
little noise, each observation's own slope really does recover its own
effect.

But notice the last row.  With $\sigma_e$ set to **zero** the Pearson
correlation is 0.906, not 1.  Noise is not the only thing separating
$s_i$ from $b_i$.  Even with no error at all,
$s_i = b_i - \bar y/(x_i-\bar x)$, and that second term explodes for
observations sitting near the mean: the largest $|s_i-b_i|$ is 25.9
overall but 0.0074 among the far half.  The irreducible part of the gap
is the centring, not the noise --- and it lives exactly where the weight
is smallest.

### What this means for the whole exercise

The identity in terms of $s_i$ holds in *any* dataset.  No model, no
heterogeneity, nothing assumed --- it would hold on pure noise.  That
also makes it empty: its weighted average is $\hat\beta$, which we
already had, so computing the $s_i$ tells us nothing new.

The version in terms of $b_i$ is the one with content, and it costs an
assumption: that $y_i=b_i x_i+\text{noise}$ describes the data.  Only
then does the coefficient estimate a particular mixture of *structural*
effects, and only then do Angrist's $\hat d(x)(1-\hat d(x))$ weighting
or Słoczyński's ATT/ATU split say anything.  Neither statement can be
made in the language of $s_i$.

So the useful part is neither slope.  It is the **weights**, which are
computable exactly, need no model, and are the only piece of this that
can be reported in a paper: they say which observations the estimate is
about, whatever the effects turn out to be.

## The rule

The panel chapter met two weightings: by $T_i$ when a panel is
unbalanced, and by variance when a regression mixes within and between
variation.  Neither is a panel-specific quirk.  Both follow from one
rule.

Let $V$ be the regressor of interest and $\tilde V$ be $V$ after
residualising on every other regressor in the model.  By the
Frisch--Waugh--Lovell theorem,

$$
\hat\beta = \frac{\sum_i \sum_t \tilde V_{it}\, y_{it}}
                 {\sum_i \sum_t \tilde V_{it}^{2}} .
$$

Be careful what $\tilde V_{it}$ is.  A single observation has no
variation of its own --- variation belongs to a collection.  What it has
is a **deviation**: how far its regressor sits from what the rest of the
model predicts for it.  With only an intercept that is distance from the
grand mean; with unit effects it is distance from the unit's own mean.
Which mean gets subtracted is the single most consequential detail in
this section, and we return to it.

### Two ways to read the coefficient

The formula supports two readings.  They describe one estimator and
differ in what is being weighted.

$$
\hat\beta = \sum_i \omega_i\, y_i , \quad
\omega_i = \frac{\tilde V_i}{\sum_j \tilde V_j^2}
\qquad\text{versus}\qquad
\hat\beta = \frac{\sum_i w_i\, \beta_i}{\sum_i w_i} , \quad
w_i = \tilde V_i V_i .
$$

| | Weights on outcomes, $\omega_i$ | Weights on effects, $w_i$ |
|---|---|---|
| Weights what | the observed $y_i$ | that observation's own effect $\beta_i$ |
| Sign | signed, sums to zero | non-negative |
| Computable | yes, $y$ is observed | no, the $\beta_i$ are never seen |
| Holds when | always, it is algebra | after writing effects into $y$; the per-unit reading also needs saturation |
| Answers | which rows drive this number, and does any enter with the wrong sign | is this the effect I meant, and whose effect is it mostly |

Note how $w_i$ relates to the weight used earlier for $s_i$.  There the
weight was $d_i^2=(x_i-\bar x)^2$, because $s_i$ was the object being
averaged; here it is $w_i=\tilde V_i V_i$, because $\beta_i$ is.  They
are different numbers observation by observation and coincide only when
$x$ is centred, though they always sum to the same total, since
$\sum_i d_i x_i=\sum_i d_i^2$.  Different object weighted, different
weight.

Default to the second.  Nearly every question that motivates the topic
--- is my coefficient the average effect, whose effect does it report,
why do smaller groups get larger weights --- is a question about the
estimand.  Reach for the first when overlap is in doubt and you need to
see whether observations enter negatively, which squares cannot show.
Both are developed at length in
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html).

### Where the square comes from

The square in $w_i$ is derived, not assumed.  Substituting
$y_{it}=\beta_i V_{it}+\ldots$ into the numerator gives
$\sum_i \beta_i \sum_t \tilde V_{it}V_{it}$, and over the whole sample
$\sum \tilde V V=\sum \tilde V^2$ because $\tilde V$ is orthogonal to the
fitted part of $V$.  Whether that holds *unit by unit* --- which is what
licenses reading $\sum_t \tilde V_{it}^2$ as unit $i$'s weight ---
depends on the residualising being done within the unit.  Demeaning
delivers it by direct expansion,
$\sum_t (V_{it}-\bar V_i)V_{it}=\sum_t V_{it}^2 - T\bar V_i^{\,2}
=\sum_t (V_{it}-\bar V_i)^2$, while a coarser control matches only in
total.  Hazlett and Shinkre make this the centre of their treatment; the
Angrist row below assumes the saturated case.

### When the weight simplifies

In general $\sum_t \tilde V_{it}^2$ will not factor into anything
tidier, since $\tilde V_{it}$ carries both subscripts even when $V$ does
not.  What changes from model to model is whether it happens to
simplify.  The table borrows the panel chapter's notation --- $x_{it}$
varying over units and time, $z_i$ over units only, $T$ periods per
unit --- because that is where the cases are easiest to name; nothing
in the rule depends on having a panel.

| Model | $\sum_t \tilde V_{it}^2$ simplifies to |
|---|---|
| $z_i$ alone, nothing else varying within the unit | $T_i\,\tilde z_i^2$ |
| a time-varying $x_{it}$, unit effects removed | $(T-1)\,\widehat{\text{Var}}_i(x)$ |
| binary $d$ with saturated controls | $n_x\,\hat d(x)\,(1-\hat d(x))$ |

Add a time-varying $x_{it}$ alongside $z_i$ in the first model and the
simplification is gone: residualising $z_i$ on something that moves
within the unit makes $\tilde z$ move within the unit too.  The rule
survives, the shortcut does not, and the $T_i$ form is the one-regressor
special case rather than the general one.  The last row is Angrist
([1998](https://doi.org/10.2307/2998558)), whose strata weights
$\hat d(x)(1-\hat d(x))$ are largest where treatment is closest to a
coin flip.  Same rule, different regressor.

## Angrist and Słoczyński, exactly

Both results are usually met as free-standing facts about treatment
effects.  They are the rule above, read at two different granularities.

### Angrist (1998): the rule with a binary regressor

Let $d$ be binary and let the controls be **saturated** --- a full set
of dummies for each covariate stratum $x$.  Residualising $d$ on those
dummies subtracts each stratum's own treated share, so
$\tilde d_i = d_i - \hat d(x)$ for observation $i$ in stratum $x$.  The
stratum's total effect weight is then

$$
\sum_{i\in x}\tilde d_i d_i
=\sum_{i\in x} d_i^2 - \hat d(x)\!\sum_{i\in x} d_i
= n_x\hat d(x) - n_x \hat d(x)^2
= n_x\,\hat d(x)\bigl(1-\hat d(x)\bigr),
$$

which is Angrist's weight.  Nothing was added: it is
$\sum \tilde V V$ with a binary $V$, where $d_i^2=d_i$ makes the sum
collapse.  The maximum at $\hat d(x)=\tfrac12$ is just the fact that a
binary variable has most variance when it is a coin flip.

Note also what plays the role of $s_i$ here.  A single observation with
a binary regressor has no slope --- you need one treated and one control
to form a difference.  So the smallest unit carrying an effect is the
*stratum*, and $s_i$ becomes the stratum difference in means.  The
granularity of the decomposition is set by the regressor, not chosen.

```{r}
set.seed(21)
n <- 400000
X <- sample(1:5, n, TRUE)

compare <- function(px) {
  d  <- rbinom(n, 1, px[X]); ta <- c(0, 1, 2, 3, 4)[X]   # effect varies by stratum
  yy <- ta * d + X + rnorm(n)
  dt <- resid(lm(d ~ factor(X)))                          # residualised treatment
  nx <- as.vector(table(X)); dh <- as.vector(tapply(d, X, mean))
  DIM <- as.vector(tapply(yy[d == 1], X[d == 1], mean) -
                   tapply(yy[d == 0], X[d == 0], mean))
  rho <- mean(d); ATT <- weighted.mean(ta, d); ATU <- weighted.mean(ta, 1 - d)
  c(`treated share rho`      = rho,
    `max |ours - Angrist|`   = max(abs(as.vector(tapply(dt * d, X, sum)) - nx*dh*(1-dh))),
    `OLS, saturated`         = unname(coef(lm(yy ~ d + factor(X)))[["d"]]),
    `Angrist weighted avg`   = sum(nx*dh*(1-dh) * DIM) / sum(nx*dh*(1-dh)),
    `Sloczynski, raw rho`    = (1 - rho) * ATT + rho * ATU,
    `ATE`                    = mean(ta))
}
res <- cbind(`near balanced` = compare(c(.08, .20, .50, .75, .93)),
             `few treated`   = compare(c(.02, .04, .08, .12, .20)))
cat(sprintf("max |our weight - Angrist's| : %.1e , %.1e\n\n", res[2, 1], res[2, 2]))
round(res[-2, ], 4)
```

The two weight vectors agree to within $6\times10^{-10}$ in both
designs, and Angrist's weighted average reproduces the OLS coefficient
exactly.  In the near-balanced design the coefficient is 2.012 against
an ATE of 1.998 --- close, because balance is where the weighting does
least.  Push the treated share down to 0.09 and the coefficient goes to
2.874 against the same ATE of 1.998.  Nothing about the effects changed;
only the treated share did.

### Słoczyński (2022): the same weights, aggregated to two groups

Angrist sorts by covariate stratum.  Słoczyński
([2022](https://doi.org/10.1162/rest_a_00953)) sorts by treatment status
instead, and collapses everything into two numbers: the effect on the
treated and the effect on the untreated.  Writing $\rho$ for the treated
share,

$$
\hat\beta \;\to\; (1-\rho)\,\text{ATT} \;+\; \rho\,\text{ATU},
\qquad\text{while}\qquad
\text{ATE} \;=\; \rho\,\text{ATT} + (1-\rho)\,\text{ATU} .
$$

The same two quantities with the weights **swapped**.  That is the whole
content of "smaller groups get larger weights": at $\rho=0.1$ the
coefficient puts $0.9$ on the effect for the treated tenth, while the
ATE puts $0.1$ there.

One caveat, and it matters.  The clean form above with the raw share
$\rho$ is exact in the *no-covariate* case.  With covariates the weights
are covariate-adjusted, so $\rho$ is only an approximation --- visible
in the table above, where the raw-$\rho$ form gives 2.029 against an
exact 2.012, and 2.858 against an exact 2.874.  Close enough to reason
with, not close enough to quote.

### Where they sit relative to each other

| | Sorts observations by | Smallest unit with an effect | Exact when |
|---|---|---|---|
| Our rule | nothing --- observation level | the observation, $s_i$ | always (algebra) |
| Angrist (1998) | covariate stratum | the stratum, a difference in means | controls are saturated |
| Słoczyński (2022) | treated vs untreated | the group, ATT and ATU | no covariates, else approximate |

Reading down the last column is the useful part.  The observation-level
identity always holds and says little.  Angrist's needs saturation and
buys an interpretable weight.  Słoczyński's needs more and buys the
sharpest statement of all --- that the coefficient and the ATE are the
same two numbers with the weights reversed.

Hazlett and Shinkre ([2024](https://arxiv.org/abs/2403.03299)) sit
underneath all three: they show these weights arise from, and
characterise, the bias caused by fitting a constant effect when effects
are heterogeneous --- and argue for estimators that tolerate
heterogeneity rather than for diagnosing the weights after the fact.

## When the weighting changes the answer

The weighting always happens.  It changes the answer only when the
effect is heterogeneous *and* the heterogeneity lines up with the
weight.  The claim to check is that a regression does not report *the*
effect but a weighted average of the effects, and that these are
different numbers.

The individual effects are never observed, so we plant them: give every
unit its own slope $\beta_i$, generate $y$ from those slopes, and see
which average of the planted slopes the estimate lands on.  In the next
chunk only one estimator is run; anything else reported is not an
estimate but the planted slopes averaged some way.  Fixed effects is
used because in a panel the natural place for heterogeneity is across
units, and it is still OLS --- OLS with the unit dummies partialled out.

Start with the extreme case.  Half the units barely move and have a true
slope of zero; the other half move a great deal and have a true slope of
two.  The average unit's slope is one.

```{r}
set.seed(1)
half <- N / 2
g <- data.frame(id   = rep(1:N, each = T),
                move = rep(c(rep(0.2, half), rep(3, half)), each = T),
                b    = rep(c(rep(0,   half), rep(2, half)), each = T))
g$x <- rnorm(N * T, 0, g$move)
g$y <- g$b * g$x + rep(rnorm(N), each = T) + rnorm(N * T)

wg <- tapply(resid(feols(x ~ 1 | id, g))^2, g$id, sum)
round(cbind(value = c(
  `average of the two true slopes`      = mean(c(0, 2)),
  `FE estimate`                         = coef(feols(y ~ x | id, g))[["x"]],
  `weight held by the steady half (%)`  = 100 * sum(wg[1:half]) / sum(wg))), 3)
```

The average slope is one and the regression reports 1.98, because the
five hundred steady units hold under half of one percent of the weight.
Their six observations sit almost on top of their own unit mean ---
this regression carries unit effects, so that is the relevant deviation
--- and squaring puts them nowhere in the sum.  Nothing has gone wrong
in the arithmetic.  The units with a slope of zero were never really in
the regression.

### Which mean, and why it matters

That result is about the *within-unit* deviation.  Plain OLS uses
distance from the grand mean instead, and the two coincided above only
because both halves were built around zero.  Separate them and the
algebra says exactly what happens.

Take two equally sized groups with means $\mu_g$, within-group variances
$\sigma_g^2$, slopes $\beta_g$ and group mean outcomes $\bar y_g$.  Let
$G$ be the group label, equally likely to be 1 or 2.  Two laws do the
work.

The **law of total covariance** splits the numerator:

$$
\text{Cov}(x,y)=E\!\left[\text{Cov}(x,y\mid G)\right]
              + \text{Cov}\!\left(E[x\mid G],\,E[y\mid G]\right).
$$

Inside group $g$ we have $\text{Cov}(x,y\mid G=g)=\beta_g\sigma_g^2$, so
the first term averages to
$\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)$.  For the second,
$E[x\mid G]$ and $E[y\mid G]$ are two-point variables taking
$(\mu_1,\mu_2)$ and $(\bar y_1,\bar y_2)$ with probability one half
each, and for any such pair
$\text{Cov}(a,b)=\tfrac14(a_1-a_2)(b_1-b_2)$.  That gives
$\tfrac14\Delta\mu\,\Delta\bar y$.

The **law of total variance** is the same statement with $y$ replaced by
$x$:

$$
\text{Var}(x)=E\!\left[\text{Var}(x\mid G)\right]+\text{Var}\!\left(E[x\mid G]\right)
=\tfrac12(\sigma_1^2+\sigma_2^2)+\tfrac14\Delta\mu^2 .
$$

Dividing one by the other,

$$
\text{plim}\;\hat\beta_{OLS}
= \frac{\tfrac12\!\left(\beta_1\sigma_1^2+\beta_2\sigma_2^2\right)
        + \tfrac14\,\Delta\mu\,\Delta\bar y}
       {\tfrac12\!\left(\sigma_1^2+\sigma_2^2\right)
        + \tfrac14\,\Delta\mu^{2}} ,
$$

with $\Delta\mu=\mu_1-\mu_2$ and $\Delta\bar y=\bar y_1-\bar y_2$.
Split that into the two slopes it mixes,

$$
\beta_{within}=\frac{\beta_1\sigma_1^2+\beta_2\sigma_2^2}{\sigma_1^2+\sigma_2^2},
\qquad
\beta_{between}=\frac{\Delta\bar y}{\Delta\mu},
$$

and the coefficient is a genuine convex combination of them, weighted by
the within variance $\tfrac12(\sigma_1^2+\sigma_2^2)$ and the between
variance $\tfrac14\Delta\mu^2$, which sum to the total variance.

Two features are worth pausing on.  $\beta_{within}$ is itself a
variance-weighted average of the two slopes --- the same $\sigma^2$
weighting as everywhere else in this section, one level up.  And
$\beta_{between}$ is a two-point slope, rise over run between the two
group centres, so it can take any value at all regardless of what
happens inside either group.  The $\tfrac14$ is the only piece specific
to equal group sizes; with shares $p$ and $1-p$ it becomes $p(1-p)$,
which is Angrist's weight again.

Everything follows without running anything.  Fixed effects estimates
$\beta_{within}$, here $(0\cdot 0.04 + 2\cdot 9)/9.04 = 1.991$, which
involves no $\mu_g$ --- so parking a group anywhere cannot move it.
Plain OLS also carries $\beta_{between}$, which is zero in this design
because the steady slope and both intercepts are zero, leaving the two
group means at the same height.  As $\Delta\mu$ grows, the between
weight $\tfrac14\Delta\mu^2$ swamps the fixed within weight of $4.52$
and the coefficient is dragged from $1.991$ toward zero like
$9/(4.52+\Delta\mu^2/4)$, the numerator being
$\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)=9$.  The simulation is a
check on that, not the source of it.

```{r}
b_within <- (0 * 0.2^2 + 2 * 3^2) / (0.2^2 + 3^2)   # variance-weighted slope
V_within <- 0.5 * (0.2^2 + 3^2)                     # within variance
predicted <- function(shift)                        # beta_between = 0 here
  V_within * b_within / (V_within + 0.25 * shift^2)

park <- function(shift) {
  set.seed(1)
  x <- rnorm(N * T, 0, g$move) + rep(c(rep(shift, half), rep(0, half)), each = T)
  y <- g$b * x + rep(rnorm(N), each = T) + rnorm(N * T)
  d <- data.frame(id = g$id, x = x, y = y)
  c(`FE (theory 1.991)` = coef(feols(y ~ x | id, d))[["x"]],
    `OLS predicted`     = predicted(shift),
    `OLS actual`        = coef(lm(y ~ x, d))[["x"]])
}

round(t(sapply(c(`shift 0` = 0, `shift 10` = 10, `shift 100` = 100), park)), 3)
```

### The same thing as a picture

The algebra says the between term takes over as the groups separate.
That is visible directly.  Left, the two groups share a centre; right,
the steady group is parked ten units away.  Dotted line is the mean of
$x$, dashed line joins the two group centres, solid line is the fit.

```{r}
#| fig-width: 9
#| fig-height: 4.2
set.seed(5)
m <- 200
draw <- function(shift, main) {
  xw <- rnorm(m, 0, 3);       yw <- 2 * xw + rnorm(m)      # wobbly: true slope 2
  xs <- rnorm(m, shift, 0.2); ys <- rnorm(m)               # steady: true slope 0
  x <- c(xw, xs); y <- c(yw, ys)
  plot(x, y, pch = 16, cex = .5, col = c(rep("#9E431855", m), rep("#0E6473", m)),
       xlab = "x", ylab = "y", main = main)
  abline(v = mean(x), lty = 3, col = "grey50")
  segments(mean(xw), mean(yw), mean(xs), mean(ys), lty = 2, lwd = 2, col = "grey30")
  points(c(mean(xw), mean(xs)), c(mean(yw), mean(ys)), pch = 21, bg = "white", cex = 1.4)
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")
  dmu <- mean(xs) - mean(xw)
  Vb  <- 0.25 * dmu^2                       # between variance
  Vw  <- 0.5 * (var(xs) + var(xw))          # within variance
  c(`slope inside wobbly` = unname(coef(lm(yw ~ xw))[2]),
    `slope inside steady` = unname(coef(lm(ys ~ xs))[2]),
    `between slope`       = unname((mean(ys) - mean(yw)) / dmu),
    `weight on between %` = 100 * Vb / (Vb + Vw),
    `OLS`                 = unname(coef(lm(y ~ x))[2]))
}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
a <- draw(0,  "groups share a centre")
b <- draw(10, "groups far apart in x")
par(op)
round(cbind(`share a centre` = a, `far apart` = b), 3)
```

Read the two columns.  The slope inside the wobbly cloud is about 2 in
both, and the slope inside the steady strip is indistinguishable from
zero in both --- $0.007$ and $-0.288$, the second noisy because a
regression on an $x$ that barely varies has a standard error of roughly
$0.35$.  The groups themselves are unchanged.  What changes is the
weight on the between comparison: 0.03% when the centres coincide, 85%
when they are ten apart.  So the fit lands on 2.01 in the first panel
and 0.29 in the second.

Notice the between slope reported for the left panel, 0.882.  It is
meaningless there, and harmlessly so: with the centres nearly on top of
each other $\Delta\mu\approx 0$, the ratio $\Delta\bar y/\Delta\mu$
divides by almost nothing, and its weight $\tfrac14\Delta\mu^2$ is
almost nothing too.  An unstable quantity multiplied by a vanishing
weight contributes nothing.

The picture also shows why the strip has influence in the right panel
despite barely moving.  Influence is distance from the mean of $x$, the
dotted line, and the strip sits ten units from it.  What it contributes
is not its own slope but its position: it anchors one end of the dashed
line.  A group dummy removes that channel and returns the within answer.

Predicted and actual agree in every row.  Fixed effects is flat at 1.984
against a theoretical 1.991, the gap being sampling noise.  Plain OLS
runs from 1.983 to 0.004 on identical true slopes, partly because the
steady half gains weight as it leaves the grand mean, and partly because
a second channel opens that is not weighting at all: with the halves at
different levels of $x$, part of the fitted line is a comparison
*between* them.  So "the weight is how much $x$ moves" describes the
demeaned case; it is exact once unit effects are removed and false in
general.

### When it does not matter

Weighting only distorts when the effect is related to the weight.  Give
every unit its own slope, let units differ in how much $x$ moves, and
compare two designs: in (A) the slopes are unrelated to that movement,
in (B) the slope is larger where $x$ moves more.  Both average near one.

```{r}
set.seed(7)
sdX <- runif(N, 0.2, 3)          # how much x moves, unit by unit

panel <- function(b) {           # build a panel whose unit slopes are b
  h   <- data.frame(id = rep(1:N, each = T),
                    x  = rnorm(N * T, 0, rep(sdX, each = T)))
  h$y <- rep(b, each = T) * h$x + rep(rnorm(N), each = T) + rnorm(N * T)
  h
}
compare <- function(h, b) {
  w <- tapply(resid(feols(x ~ 1 | id, h))^2, h$id, sum)   # the weights
  c(`FE estimate`            = coef(feols(y ~ x | id, h))[["x"]],
    `true slopes, plain avg` = mean(b),
    `true slopes, weighted`  = sum(w * b) / sum(w))
}

b1 <- rnorm(N, 1, 1)             # (A) slopes unrelated to movement
b2 <- 1 + 0.8 * (sdX - 1.6)      # (B) slopes larger where x moves more
h1 <- panel(b1); h2 <- panel(b2)

round(cbind(`(A) unrelated` = compare(h1, b1),
            `(B) aligned`   = compare(h2, b2)), 4)
```

In (A) all three numbers agree to within sampling noise: the weighting
is there and costs nothing.  In (B) the estimate is 1.53 while the plain
average of the true slopes is 1.01, and the weighted average, 1.54, is
what the estimator was tracking all along.  None of this is a defect ---
a unit whose $x$ barely moves genuinely carries little information about
how $y$ responds, and leaning on the units that do move is efficient.

Design (B) is the continuous version of what Słoczyński describes for a
binary treatment, worked through earlier in this chapter: the same
weighting, aggregated to two groups instead of spread over a continuum.

With homogeneous effects the weighting is harmless.  With heterogeneous
effects the question is never whether OLS is weighting, only whether we
know what it weighted.
