# Interpreting OLS

```{r}
#| include: false
library(fixest)
N <- 1000; T <- 6
```

A regression returns one number.  What that number *is* --- whose effect
it reflects, which observations produced it, whether it is the average
anyone had in mind --- is a separate question from how it is computed,
and the answer is less obvious than it looks.

This chapter collects the material.  It sits at the end because it
draws on the between-versus-within split from
[Panel Data Models](panel-data.qmd), but almost nothing in it is
specific to panels: the same rule governs a plain cross-sectional
regression, and the binary-treatment case that most of the applied
literature discusses is one line of the table below.

Two companion pieces develop the same ideas with a treatment-effects
emphasis:
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html)
and
[OLS cannot be ATE](https://xiangao.github.io/blog_book/ols-ate.html).

## Start with two points

Begin as small as possible.  Two points, and a regression line through
them.  Everyone knows what that line is: it joins them.

Now describe it in a way that will keep working when there are more
points.  Take the **centroid** $(\bar x, \bar y)$ and, for each point,
the slope of the line running from the centroid out to it:

$$
s_i \;=\; \frac{y_i - \bar y}{x_i - \bar x} .
$$

Call $s_i$ that observation's own slope.  With two points the centroid
lies on the segment joining them, so both rays *are* that segment and
both $s_i$ are the same number.

Attach a weight to each, $d_i^2=(x_i-\bar x)^2$, the squared horizontal
distance from the centroid.  We will justify that choice shortly; for
now just take the weighted average.

In the figures below the hollow circle is the centroid, the dotted
vertical line is $\bar x$, the rust rays run from the centroid to each
point with **thickness proportional to that point's weight**, and the
solid line is the fit.  Percentages label each point's share of the
total weight.

```{r}
#| fig-width: 5
#| fig-height: 4
show <- function(x, y, main = "") {
  xb <- mean(x); yb <- mean(y); d <- x - xb; s <- (y - yb) / d; w <- d^2
  plot(x, y, pch = 16, cex = 1.5, xlim = c(-0.5, 11.5), ylim = c(0, 9),
       xlab = "x", ylab = "y", main = main)
  abline(v = xb, lty = 3, col = "grey55")                      # the mean of x
  segments(xb, yb, x, y, lwd = 0.6 + 9 * w / sum(w),           # rays, thickness = weight
           col = "#9E431899")
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")                  # the fit
  points(x, y, pch = 16, cex = 1.5)
  points(xb, yb, pch = 21, bg = "white", col = "#0E6473", lwd = 2.5, cex = 1.7)
  text(x, y, labels = sprintf("%.1f%%", 100 * w / sum(w)),   # label away from centroid
       pos = ifelse(y >= yb, 3, 1), cex = .85, col = "#9E4318")
  print(round(data.frame(x, y, s_i = s, weight = w,
                         `weight %` = 100 * w / sum(w), check.names = FALSE), 3),
        row.names = FALSE)
  cat(sprintf("  centroid x = %.3f | weighted avg of s_i = %.4f | OLS = %.4f\n",
              xb, sum(w * s) / sum(w), coef(lm(y ~ x))[["x"]]))
}
show(c(1, 3), c(2, 8), "two points")
```

Both $s_i$ equal 3, the weights split evenly, and the average is 3 ---
which is the fitted slope.  Note what this means: with two points the
weights are *irrelevant*.  Any weighting of two identical numbers
returns that number.

## Add a third point

Same function, one more point, at $x=4$.

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4), c(2, 8, 5), "three points")
```

Now the three own-slopes disagree: 1.8, 9.0 and 0.0.  The fit has to
choose among them, and it does so by weight --- and the weights are
wildly uneven.  The point at $x=3$ has the steepest own-slope of the
three, 9.0, and receives 2.4% of the say, because $x=3$ sits almost on
the centroid at 2.667.  The answer, 1.2857, is the weighted average.

This is the whole mechanism.  A point close to $\bar x$ has almost no
horizontal leverage, so however dramatic its own slope, it barely moves
the line.

## Add a fourth, far out

Again the same function, now with a point at $x=9$, well to the right of
everything else.

```{r}
#| fig-width: 5
#| fig-height: 4
show(c(1, 3, 4, 9), c(2, 8, 5, 6), "four points")
```

Two things happened at once, and both are worth naming.

The new point at $x=9$ took **65% of the weight by itself**, because
weight is squared distance and it is far from centre.  And the point at
$x=4$, which had 38% of the say a moment ago, collapsed to **0.18%** ---
not because it changed, but because the centroid moved out to 4.25 and
landed almost on top of it.

So weight is not a property of an observation.  It is a property of an
observation *relative to the rest of the sample*, and adding data
anywhere rewrites it everywhere.  The fitted slope fell from 1.29 to
0.31 without a single existing point being touched.

## Add a hundred points

The same picture, at a scale where individual points can no longer be
labelled.  A hundred points scattered around a centre, rays as before,
thickness proportional to weight.

```{r}
#| fig-width: 6.5
#| fig-height: 5
set.seed(3)
n <- 100
x <- rnorm(n, 5, 2)
y <- 2 * x + rnorm(n, 0, 3)

xb <- mean(x); yb <- mean(y); w <- (x - xb)^2
o <- order(w, decreasing = TRUE)

plot(x, y, pch = 16, cex = 1, col = "grey40",
     xlab = "x", ylab = "y", main = "a hundred points")
abline(v = xb, lty = 3, col = "grey55")
segments(xb, yb, x, y, lwd = 0.3 + 10 * w / max(w), col = "#9E431870")
abline(lm(y ~ x), lwd = 2.5, col = "#0E6473")
points(x, y, pch = 16, cex = 1, col = "grey25")
points(xb, yb, pch = 21, bg = "white", col = "#0E6473", lwd = 2.5, cex = 1.8)

round(c(`weight held by the 10 points nearest the mean` = 100 * sum(w[o[91:100]]) / sum(w),
        `weight held by the 10 points furthest out`     = 100 * sum(w[o[1:10]]) / sum(w),
        `OLS`                                            = coef(lm(y ~ x))[["x"]]), 3)
```

Look at the rays rather than the points.  Near the dotted line they are
too thin to see at all; out at the edges they are thick enough to
dominate the picture.  The ten points nearest the mean hold **0.14%** of
the total weight between them.  The ten points furthest out hold
**35%** --- from a tenth of the sample.

Nothing about this is particular to having only two, three or four
points.  It was never a small-sample curiosity: the same handful of
extreme observations would run the fit in a sample of a hundred, a
thousand, or more.

## The identity behind it

None of that was a convention.  Since
$(x_i-\bar x)(y_i-\bar y)=(x_i-\bar x)^2 s_i$, the ordinary OLS formula
*is* that weighted average:

$$
\hat\beta
=\frac{\sum_i (x_i-\bar x)(y_i-\bar y)}{\sum_i (x_i-\bar x)^2}
=\frac{\sum_i (x_i-\bar x)^2\, s_i}{\sum_i (x_i-\bar x)^2}.
$$

Every observation carries a slope of its own, and the fit averages them
with weights $d_i^2$.  The two-group example later in this chapter is
this same arithmetic with three thousand points on each side instead of
two.

One consequence is worth stating before we go on.  An observation
sitting exactly at $\bar x$ has an undefined $s_i$ and zero weight ---
the wild value arrives paired with a weight that annihilates it.  The
four-point panel shows this happening in miniature.

## Do not mistake $s_i$ for $b_i$

Suppose, hypothetically, that every observation really did have its own
slope --- $y_i = b_i x_i + \text{noise}$, with $b_i$ a fixed but unknown
number specific to observation $i$.  $b_i$ is the *true* slope; $s_i$ is
our earlier point-to-centroid slope, computed from the data.

This is where care is needed.  $s_i$ is *observable* --- it is computed
from the data.  The structural $b_i$ is not.  It is tempting to treat
the first as a measurement of the second, and at the level of a single
observation that is badly wrong.

### The heterogeneous case

To see it we need both quantities at once, so we plant the truth: two
thousand observations, each given its own slope $b_i$ drawn around 1,
with a fair amount of noise on top.  Then we compute $s_i$ from the data
the ordinary way and compare the two.

```{r}
set.seed(6)
n <- 2000
b <- rnorm(n, 1, 0.5)                 # TRUE slope of each observation, unobservable
x <- rnorm(n, 0, 2)
y <- b * x + rnorm(n, 0, 2)
xb <- mean(x); w <- (x - xb)^2
s  <- (y - mean(y)) / (x - xb)        # OBSERVABLE slope to the centroid

round(cbind(value = c(
  `correlation of s_i with b_i` = cor(s, b),
  `sd of true b_i`              = sd(b),
  `sd of observed s_i`          = sd(s),
  `weighted average of s_i`     = sum(w * s) / sum(w),
  `weighted average of b_i`     = sum(w * b) / sum(w),
  `OLS estimate`                = unname(coef(lm(y ~ x))[2]))), 3)
```

The Pearson correlation is 0.019 and the spread of the observed $s_i$ is
more than two hundred times the spread of the true $b_i$, because $s_i$
divides a noisy vertical gap by a horizontal distance that is sometimes
almost zero.  An observation with an ordinary $b_i$ near one can easily
show an $s_i$ of eighteen.

Do not over-read that 0.019, though.  It is a Pearson correlation on a
variable with enormous tails, so a handful of exploded $s_i$ dominate
it; the rank correlation on the same data is far higher.  The signal is
present, and buried, which is a different thing from absent.

Yet the weighted averages agree: $0.950$ for the $s_i$, $0.992$ for the
$b_i$, and the OLS estimate is $0.950$ --- identical to the $s_i$
version by construction, and close to the $b_i$ version because the
noise averages out.

That comparison had genuine noise sitting on top of the heterogeneity.
The two are different phenomena, and it is worth seeing them apart.
Turn the noise off entirely --- let $y_i=b_i x_i$ exactly, nothing
added --- and ask how well $s_i$ tracks its own $b_i$.

```{r}
y0 <- b * x                                  # same b, same x, no noise at all
d  <- x - xb
s0 <- (y0 - mean(y0)) / d
round(c(`correlation, s_i with b_i, no noise` = cor(s0, b),
        `weighted average of s_i`             = sum(w * s0) / sum(w),
        `weighted average of b_i`              = sum(w * b) / sum(w),
        `OLS on this noiseless data`           = unname(coef(lm(y0 ~ x))[2])), 4)
```

The correlation is 0.14 --- still low, with no noise anywhere in the
data.  This is the two-, three- and four-point geometry from the start
of the chapter, applied to a whole sample instead of a handful of
points: a ray's slope reflects where *every* point sits, not only its
own, so when the $b_i$ genuinely differ, one point's ray does not
recover that point's own slope even without noise.  Noise makes it
worse, but it is not the cause of it.

Write out exactly what happens.  Let $\bar b$ be the average $b_i$ and
$C$ the sample covariance of $b$ and $x$.  Substituting $x_i=d_i+\bar x$
throughout and collecting terms,

$$
s_i \;=\; b_i \;+\; \frac{\bar x\,(b_i-\bar b) - C}{d_i}
        \;+\; \frac{e_i-\bar e}{d_i}.
$$

Set $b_i=b$ for everyone and the middle term vanishes on its own, since
$b_i-\bar b=0$ and $C=0$, leaving exactly the homogeneous formula used
below.  With $b_i$ heterogeneous, that middle term survives, and it
divides by $d_i$ exactly like the noise term does --- so it inflates
near the mean for the same geometric reason, with no noise involved at
all.  It vanishes only if $x$ happens to be centred exactly
($\bar x=0$) and $b$ happens to be uncorrelated with $x$ in this
particular sample ($C=0$); neither is guaranteed at any finite $n$.

What survives regardless is the weighted average, which comes out at
$0.9914$ either way and matches OLS exactly.  That is the weighted-
average identity established earlier in this chapter: it is pure
algebra, and needs nothing about noise or homogeneity to hold.

Put noise back and ask the harder question directly: is $s_i$ unbiased
for its own $b_i$, holding $b$ and $x$ fixed and averaging only over
$e$?  Take the expectation of the formula above,

$$
E(s_i) \;=\; b_i + \frac{\bar x(b_i-\bar b) - C}{d_i} + 0 .
$$

The noise term vanishes as always, but $\bar x$, $\bar b$ and $C$ are
fixed numbers once $b$ and $x$ are fixed --- nothing left to average
over --- so the bracket does not go anywhere.  $s_i$ is **not** unbiased
for its own $b_i$; the bias is permanent, and it has nothing to do with
noise.

Nothing about that kills the estimator, though, and here is exactly
why.  $\hat\beta$ is not merely *related* to the $s_i$ --- by the
identity that opened this chapter, $\hat\beta=\sum_i d_i^2 s_i
/\sum_i d_i^2$ exactly, and $\sum_i d_i^2$ is a fixed number that does
not depend on $e$.  So expectation passes straight through:

$$
E(\hat\beta) \;=\; \frac{\sum_i d_i^2\, E(s_i)}{\sum_i d_i^2}.
$$

$E(\hat\beta)$ is, quite literally, the $d_i^2$-weighted average of what
each $s_i$ is expected to be.  $s_i$ has not gone anywhere.

Substitute $E(s_i)=b_i+[\bar x(b_i-\bar b)-C]/d_i$, and the bias terms
do *not* cancel to zero.  Weighted by $d_i^2$ and summed, they add up to

$$
\sum_i d_i^2 \cdot \frac{\bar x(b_i-\bar b)-C}{d_i}
= \bar x \sum_i d_i(b_i-\bar b) - C\sum_i d_i
= \bar x\, nC,
$$

using $\sum_i d_i=0$ and $\sum_i d_i(b_i-\bar b)=\sum_i d_i b_i=nC$ by
the definition of $C$.  That is a specific, generally nonzero number,
not something that averages away.  What it does is combine with
$\sum_i d_i^2 b_i$ to give exactly

$$
\sum_i d_i^2\, E(s_i) = \sum_i d_i^2 b_i + n\bar x\,C
= \sum_i b_i\,d_i(d_i+\bar x) = \sum_i b_i\, d_i x_i,
$$

using $x_i=d_i+\bar x$.  Divide by $\sum_i d_i^2$, which --- as noted
above --- equals $\sum_i d_i x_i$ exactly:

$$
E(\hat\beta) \;=\; \frac{\sum_i d_i^2\, E(s_i)}{\sum_i d_i^2}
\;=\; \frac{\sum_i d_i x_i\, b_i}{\sum_i d_i x_i},
$$

exactly the effect-weighted average of $b_i$ from "Two ways to read the
coefficient" earlier in this chapter --- no approximation, no large $n$,
true at every sample size.

That is the whole answer.  Each $s_i$'s own bias is permanent and never
shrinks, however large the sample.  Weighted by $d_i^2$ and summed,
those biases do not cancel to nothing either --- they sum to a real,
nonzero quantity, $\bar x\,nC$.  What that quantity does is convert one
valid weighted average of the true $b_i$, weighted by $d_i^2$, into
another valid weighted average of the same $b_i$, weighted by
$d_i x_i$ --- the one OLS is supposed to deliver.  Individually wrong in
a specific and permanent way; combined, exactly right, because the
wrongness reshapes the weighting rather than vanishing.

### Why averaging $s_i$ works

Set the heterogeneity question aside, then --- it is handled --- and ask
only about the noise: is $s_i$ unbiased for $b$, and why does $d_i^2$ in
particular make a good weight?

Everything below is **conditional on $X$**, as usual in regression, and
that matters twice over: it makes $d_i=x_i-\bar x$ a known constant
rather than a random variable, and the weights are functions of $x$
alone, so the whole weighting story only lives in the conditional world.

With homogeneous truth $b$, $y_i-\bar y = b\,d_i + (e_i-\bar e)$, so

$$
s_i = b + \frac{e_i-\bar e}{d_i}.
$$

Rearrange and take an expectation: $b = s_i - (e_i-\bar e)/d_i$, and
since $E(e_i-\bar e\mid X)=0$, $E(s_i\mid X)=b$.  So $s_i$ is unbiased,
and any weighted average of the $s_i$ --- equal weights, $d_i^2$,
anything --- is unbiased for $b$ too.  What the weights buy is not
correctness but precision: some $s_i$ are far noisier than others, and
$d_i^2$ happens to be the weighting that makes the average as precise as
possible.  Since
$\text{Var}(e_i-\bar e)=\sigma^2-2\sigma^2/n+\sigma^2/n=\sigma^2(1-1/n)$,

$$
\text{Var}(s_i)\;=\;\frac{\sigma^2\,(1-1/n)}{d_i^{\,2}}
\;\approx\;\frac{\sigma^2}{d_i^{\,2}},
\qquad\text{so}\qquad
d_i^{\,2} \;\propto\; \frac{1}{\text{Var}(s_i)} .
$$

So an observation's weight is the reciprocal of how noisy its own slope
estimate is.  The $s_i$ that look absurd are absurd because their $d_i$
is small, and a small $d_i$ is exactly what the weight penalises.

The cancellation behind that weight is algebraic, not approximate:
$d_i^2\, s_i = d_i^2 \cdot (y_i-\bar y)/d_i = d_i(y_i-\bar y)$, true in
any sample, which is what makes OLS the inverse-variance weighted
average of $n$ unbiased estimates of the slope.

This is the step that explains why averaging $s_i$ recovers $b$, not
merely that it is unbiased.  Because $s_i$ is unbiased,
$\text{Var}(s_i)$ is exactly the average squared distance of $s_i$ from
$b$ --- large variance means $s_i$ typically lands far away, small
variance means it typically lands close.  Since $\text{Var}(s_i)$ is
proportional to $1/d_i^{\,2}$, weighting by $d_i^2$ hands almost all the
say to observations whose $s_i$ reliably sit near $b$, and next to none
to observations whose $s_i$ wander far from it.

### What this means for the whole exercise

Writing $s_i$ out in full is what showed the individual biases
reshaping the weighting rather than cancelling --- an argument that
needed the $s_i$ algebra, not just a claim about $b_i$.  Writing effects
into $y_i=b_i x_i+\text{noise}$ is what gives the coefficient content: a
specific mixture of structural effects, the same mixture Angrist's
$\hat d(x)(1-\hat d(x))$ weighting and Słoczyński's ATT/ATU split name
one level of aggregation up.

The **weights** are the part usable without knowing any $b_i$:
computable exactly, and the only piece of this that belongs in a paper,
since they say which observations the estimate is about whatever the
effects turn out to be.

## The rule

The panel chapter met two weightings: by $T_i$ when a panel is
unbalanced, and by variance when a regression mixes within and between
variation.  Neither is a panel-specific quirk.  Both follow from one
rule.

Let $V$ be the regressor of interest and $\tilde V$ be $V$ after
residualising on every other regressor in the model.  By the
Frisch--Waugh--Lovell theorem,

$$
\hat\beta = \frac{\sum_i \tilde V_i\, y_i}{\sum_i \tilde V_i^{2}} .
$$

Be careful what $\tilde V_i$ is.  A single observation has no
variation of its own --- variation belongs to a collection.  What it has
is a **deviation**: how far its regressor sits from what the rest of the
model predicts for it.  With only an intercept that is distance from the
grand mean; with unit effects --- as in a panel, where several
observations share a unit --- it is distance from the unit's own mean.
Which mean gets subtracted is the single most consequential detail in
this section, and we return to it.

### Two ways to read the coefficient

The formula supports two readings.  They describe one estimator and
differ in what is being weighted.

$$
\hat\beta = \sum_i \omega_i\, y_i , \quad
\omega_i = \frac{\tilde V_i}{\sum_j \tilde V_j^2}
\qquad\text{versus}\qquad
\hat\beta = \frac{\sum_i w_i\, \beta_i}{\sum_i w_i} , \quad
w_i = \tilde V_i V_i .
$$

| | Weights on outcomes, $\omega_i$ | Weights on effects, $w_i$ |
|---|---|---|
| Weights what | the observed $y_i$ | that observation's own effect $\beta_i$ |
| Sign | signed, sums to zero | non-negative |
| Computable | yes, $y$ is observed | no, the $\beta_i$ are never seen |
| Holds when | always, it is algebra | after writing effects into $y$; the per-unit reading also needs saturation |
| Answers | which rows drive this number, and does any enter with the wrong sign | is this the effect I meant, and whose effect is it mostly |

Note how $w_i$ relates to the weight used earlier for $s_i$.  There the
weight was $d_i^2=(x_i-\bar x)^2$, because $s_i$ was the object being
averaged; here it is $w_i=\tilde V_i V_i$, because $\beta_i$ is.  They
are different numbers observation by observation and coincide only when
$x$ is centred, though they always sum to the same total, since
$\sum_i d_i x_i=\sum_i d_i^2$.  Different object weighted, different
weight.

Default to the second.  Nearly every question that motivates the topic
--- is my coefficient the average effect, whose effect does it report,
why do smaller groups get larger weights --- is a question about the
estimand.  Reach for the first when overlap is in doubt and you need to
see whether observations enter negatively, which squares cannot show.
Both are developed at length in
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html).

### Where the square comes from

The square in $w_i$ is derived, not assumed.  Substituting
$y_i=\beta_i V_i+\ldots$ into the numerator gives
$\sum_i \beta_i \tilde V_i V_i$, and over the whole sample
$\sum \tilde V V=\sum \tilde V^2$ because $\tilde V$ is orthogonal to the
fitted part of $V$.

In a panel, several observations share a unit, indexed by $t$ within
unit $i$, and it is natural to ask whether the same cancellation holds
*within a single unit* --- summed only over that unit's own $t$'s ---
rather than only over the whole sample.  Write $V_{it}$ for the
observation and $\bar V_i$ for the unit's own mean.  Demeaning delivers
the within-unit version by direct expansion,
$\sum_t (V_{it}-\bar V_i)V_{it}=\sum_t V_{it}^2 - T\bar V_i^{\,2}
=\sum_t (V_{it}-\bar V_i)^2$, while a coarser control matches only in
total, over the whole sample, not unit by unit.  Hazlett and Shinkre
make this the centre of their treatment; the Angrist row below assumes
the saturated case.

### When the weight simplifies

In general the within-unit sum $\sum_t \tilde V_{it}^2$ will not factor
into anything tidier, since $\tilde V_{it}$ carries both subscripts even
when $V$ does not.  What changes from model to model is whether it
happens to simplify.  The table is stated for a panel --- $x_{it}$
varying over units and time, $z_i$ over units only, $T$ periods per
unit --- because that is where the cases are easiest to name; nothing
in the rule depends on having a panel.

| Model | $\sum_t \tilde V_{it}^2$ simplifies to |
|---|---|
| $z_i$ alone, nothing else varying within the unit | $T_i\,\tilde z_i^2$ |
| a time-varying $x_{it}$, unit effects removed | $(T-1)\,\widehat{\text{Var}}_i(x)$ |
| binary $d$ with saturated controls | $n_x\,\hat d(x)\,(1-\hat d(x))$ |

Add a time-varying $x_{it}$ alongside $z_i$ in the first model and the
simplification is gone: residualising $z_i$ on something that moves
within the unit makes $\tilde z$ move within the unit too.  The rule
survives, the shortcut does not, and the $T_i$ form is the one-regressor
special case rather than the general one.  The last row is Angrist
([1998](https://doi.org/10.2307/2998558)), whose strata weights
$\hat d(x)(1-\hat d(x))$ are largest where treatment is closest to a
coin flip.  Same rule, different regressor.

## Angrist and Słoczyński, exactly

Both results are usually met as free-standing facts about treatment
effects.  They are the rule above, read at two different granularities.

### Angrist (1998): the rule with a binary regressor

Let $d$ be binary and let the controls be **saturated** --- a full set
of dummies for each covariate stratum $x$.  Residualising $d$ on those
dummies subtracts each stratum's own treated share, so
$\tilde d_i = d_i - \hat d(x)$ for observation $i$ in stratum $x$.  The
stratum's total effect weight is then

$$
\sum_{i\in x}\tilde d_i d_i
=\sum_{i\in x} d_i^2 - \hat d(x)\!\sum_{i\in x} d_i
= n_x\hat d(x) - n_x \hat d(x)^2
= n_x\,\hat d(x)\bigl(1-\hat d(x)\bigr),
$$

which is Angrist's weight.  Nothing was added: it is
$\sum \tilde V V$ with a binary $V$, where $d_i^2=d_i$ makes the sum
collapse.  The maximum at $\hat d(x)=\tfrac12$ is just the fact that a
binary variable has most variance when it is a coin flip.

Note also what plays the role of $s_i$ here.  A single observation with
a binary regressor has no slope --- you need one treated and one control
to form a difference.  So the smallest unit carrying an effect is the
*stratum*, and $s_i$ becomes the stratum difference in means.  The
granularity of the decomposition is set by the regressor, not chosen.

The check below builds five covariate strata with different treated
shares and a different effect in each, then computes the stratum weights
twice --- once as $\sum \tilde d_i d_i$, once as Angrist's
$n_x\hat d(1-\hat d)$ --- and compares both to the OLS coefficient and
to Słoczyński's form.  It runs the whole thing under two treated shares,
one near a half and one near a tenth.

```{r}
set.seed(21)
n <- 400000
X <- sample(1:5, n, TRUE)

compare <- function(px) {
  d  <- rbinom(n, 1, px[X]); ta <- c(0, 1, 2, 3, 4)[X]   # effect varies by stratum
  yy <- ta * d + X + rnorm(n)
  dt <- resid(lm(d ~ factor(X)))                          # residualised treatment
  nx <- as.vector(table(X)); dh <- as.vector(tapply(d, X, mean))
  DIM <- as.vector(tapply(yy[d == 1], X[d == 1], mean) -
                   tapply(yy[d == 0], X[d == 0], mean))
  rho <- mean(d); ATT <- weighted.mean(ta, d); ATU <- weighted.mean(ta, 1 - d)
  c(`treated share rho`      = rho,
    `max |ours - Angrist|`   = max(abs(as.vector(tapply(dt * d, X, sum)) - nx*dh*(1-dh))),
    `OLS, saturated`         = unname(coef(lm(yy ~ d + factor(X)))[["d"]]),
    `Angrist weighted avg`   = sum(nx*dh*(1-dh) * DIM) / sum(nx*dh*(1-dh)),
    `Sloczynski, raw rho`    = (1 - rho) * ATT + rho * ATU,
    `ATE`                    = mean(ta))
}
res <- cbind(`near balanced` = compare(c(.08, .20, .50, .75, .93)),
             `few treated`   = compare(c(.02, .04, .08, .12, .20)))
cat(sprintf("max |our weight - Angrist's| : %.1e , %.1e\n\n", res[2, 1], res[2, 2]))
round(res[-2, ], 4)
```

The two weight vectors agree to within $6\times10^{-10}$ in both
designs, and Angrist's weighted average reproduces the OLS coefficient
exactly.  In the near-balanced design the coefficient is 2.012 against
an ATE of 1.998 --- close, because balance is where the weighting does
least.  Push the treated share down to 0.09 and the coefficient goes to
2.874 against the same ATE of 1.998.  Nothing about the effects changed;
only the treated share did.

### Słoczyński (2022): the same weights, aggregated to two groups

Angrist sorts by covariate stratum.  Słoczyński
([2022](https://doi.org/10.1162/rest_a_00953)) sorts by treatment status
instead, and collapses everything into two numbers: the effect on the
treated and the effect on the untreated.  Writing $\rho$ for the treated
share,

$$
\hat\beta \;\to\; (1-\rho)\,\text{ATT} \;+\; \rho\,\text{ATU},
\qquad\text{while}\qquad
\text{ATE} \;=\; \rho\,\text{ATT} + (1-\rho)\,\text{ATU} .
$$

The same two quantities with the weights **swapped**.  That is the whole
content of "smaller groups get larger weights": at $\rho=0.1$ the
coefficient puts $0.9$ on the effect for the treated tenth, while the
ATE puts $0.1$ there.

One caveat, and it matters.  The clean form above with the raw share
$\rho$ is exact in the *no-covariate* case.  With covariates the weights
are covariate-adjusted, so $\rho$ is only an approximation --- visible
in the table above, where the raw-$\rho$ form gives 2.029 against an
exact 2.012, and 2.858 against an exact 2.874.  Close enough to reason
with, not close enough to quote.

### Where they sit relative to each other

| | Sorts observations by | Smallest unit with an effect | Exact when |
|---|---|---|---|
| Our rule | nothing --- observation level | the observation, $s_i$ | always (algebra) |
| Angrist (1998) | covariate stratum | the stratum, a difference in means | controls are saturated |
| Słoczyński (2022) | treated vs untreated | the group, ATT and ATU | no covariates, else approximate |

Reading down the last column is the useful part.  The observation-level
identity always holds and says little.  Angrist's needs saturation and
buys an interpretable weight.  Słoczyński's needs more and buys the
sharpest statement of all --- that the coefficient and the ATE are the
same two numbers with the weights reversed.

Hazlett and Shinkre ([2024](https://arxiv.org/abs/2403.03299)) sit
underneath all three: they show these weights arise from, and
characterise, the bias caused by fitting a constant effect when effects
are heterogeneous --- and argue for estimators that tolerate
heterogeneity rather than for diagnosing the weights after the fact.

## When the weighting changes the answer

The weighting always happens.  It changes the answer only when the
effect is heterogeneous *and* the heterogeneity lines up with the
weight.  The claim to check is that a regression does not report *the*
effect but a weighted average of the effects, and that these are
different numbers.

The individual effects are never observed, so we plant them: give every
unit its own slope $\beta_i$, generate $y$ from those slopes, and see
which average of the planted slopes the estimate lands on.  In the next
chunk only one estimator is run; anything else reported is not an
estimate but the planted slopes averaged some way.  Fixed effects is
used because in a panel the natural place for heterogeneity is across
units, and it is still OLS --- OLS with the unit dummies partialled out.

Start with the extreme case.  Half the units barely move and have a true
slope of zero; the other half move a great deal and have a true slope of
two.  The average unit's slope is one.

```{r}
set.seed(1)
half <- N / 2
g <- data.frame(id   = rep(1:N, each = T),
                move = rep(c(rep(0.2, half), rep(3, half)), each = T),
                b    = rep(c(rep(0,   half), rep(2, half)), each = T))
g$x <- rnorm(N * T, 0, g$move)
g$y <- g$b * g$x + rep(rnorm(N), each = T) + rnorm(N * T)

wg <- tapply(resid(feols(x ~ 1 | id, g))^2, g$id, sum)
round(cbind(value = c(
  `average of the two true slopes`      = mean(c(0, 2)),
  `FE estimate`                         = coef(feols(y ~ x | id, g))[["x"]],
  `weight held by the steady half (%)`  = 100 * sum(wg[1:half]) / sum(wg))), 3)
```

The average slope is one and the regression reports 1.98, because the
five hundred steady units hold under half of one percent of the weight.
Their six observations sit almost on top of their own unit mean ---
this regression carries unit effects, so that is the relevant deviation
--- and squaring puts them nowhere in the sum.  Nothing has gone wrong
in the arithmetic.  The units with a slope of zero were never really in
the regression.

### Which mean, and why it matters

That result is about the *within-unit* deviation.  Plain OLS uses
distance from the grand mean instead, and the two coincided above only
because both halves were built around zero.  Separate them and the
algebra says exactly what happens.

Take two equally sized groups with means $\mu_g$, within-group variances
$\sigma_g^2$, slopes $\beta_g$ and group mean outcomes $\bar y_g$.  Let
$G$ be the group label, equally likely to be 1 or 2.  Two laws do the
work.

The **law of total covariance** splits the numerator:

$$
\text{Cov}(x,y)=E\!\left[\text{Cov}(x,y\mid G)\right]
              + \text{Cov}\!\left(E[x\mid G],\,E[y\mid G]\right).
$$

Inside group $g$ we have $\text{Cov}(x,y\mid G=g)=\beta_g\sigma_g^2$, so
the first term averages to
$\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)$.  For the second,
$E[x\mid G]$ and $E[y\mid G]$ are two-point variables taking
$(\mu_1,\mu_2)$ and $(\bar y_1,\bar y_2)$ with probability one half
each, and for any such pair
$\text{Cov}(a,b)=\tfrac14(a_1-a_2)(b_1-b_2)$.  That gives
$\tfrac14\Delta\mu\,\Delta\bar y$.

The **law of total variance** is the same statement with $y$ replaced by
$x$:

$$
\text{Var}(x)=E\!\left[\text{Var}(x\mid G)\right]+\text{Var}\!\left(E[x\mid G]\right)
=\tfrac12(\sigma_1^2+\sigma_2^2)+\tfrac14\Delta\mu^2 .
$$

Dividing one by the other,

$$
\text{plim}\;\hat\beta_{OLS}
= \frac{\tfrac12\!\left(\beta_1\sigma_1^2+\beta_2\sigma_2^2\right)
        + \tfrac14\,\Delta\mu\,\Delta\bar y}
       {\tfrac12\!\left(\sigma_1^2+\sigma_2^2\right)
        + \tfrac14\,\Delta\mu^{2}} ,
$$

with $\Delta\mu=\mu_1-\mu_2$ and $\Delta\bar y=\bar y_1-\bar y_2$.
Split that into the two slopes it mixes,

$$
\beta_{within}=\frac{\beta_1\sigma_1^2+\beta_2\sigma_2^2}{\sigma_1^2+\sigma_2^2},
\qquad
\beta_{between}=\frac{\Delta\bar y}{\Delta\mu},
$$

and the coefficient is a genuine convex combination of them, weighted by
the within variance $\tfrac12(\sigma_1^2+\sigma_2^2)$ and the between
variance $\tfrac14\Delta\mu^2$, which sum to the total variance.

Two features are worth pausing on.  $\beta_{within}$ is itself a
variance-weighted average of the two slopes --- the same $\sigma^2$
weighting as everywhere else in this section, one level up.  And
$\beta_{between}$ is a two-point slope, rise over run between the two
group centres, so it can take any value at all regardless of what
happens inside either group.  The $\tfrac14$ is the only piece specific
to equal group sizes; with shares $p$ and $1-p$ it becomes $p(1-p)$,
which is Angrist's weight again.

Everything follows without running anything.  Fixed effects estimates
$\beta_{within}$, here $(0\cdot 0.04 + 2\cdot 9)/9.04 = 1.991$, which
involves no $\mu_g$ --- so parking a group anywhere cannot move it.
Plain OLS also carries $\beta_{between}$, which is zero in this design
because the steady slope and both intercepts are zero, leaving the two
group means at the same height.  As $\Delta\mu$ grows, the between
weight $\tfrac14\Delta\mu^2$ swamps the fixed within weight of $4.52$
and the coefficient is dragged from $1.991$ toward zero like
$9/(4.52+\Delta\mu^2/4)$, the numerator being
$\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)=9$.

The code below evaluates that expression, then takes the same data and
shifts the steady group to three different positions --- on top of the
other group, ten units away, a hundred units away --- reporting the
fixed-effects estimate, the predicted OLS value and the actual one side
by side.  The simulation is a check on the algebra, not the source of
it.

```{r}
b_within <- (0 * 0.2^2 + 2 * 3^2) / (0.2^2 + 3^2)   # variance-weighted slope
V_within <- 0.5 * (0.2^2 + 3^2)                     # within variance
predicted <- function(shift)                        # beta_between = 0 here
  V_within * b_within / (V_within + 0.25 * shift^2)

park <- function(shift) {
  set.seed(1)
  x <- rnorm(N * T, 0, g$move) + rep(c(rep(shift, half), rep(0, half)), each = T)
  y <- g$b * x + rep(rnorm(N), each = T) + rnorm(N * T)
  d <- data.frame(id = g$id, x = x, y = y)
  c(`FE (theory 1.991)` = coef(feols(y ~ x | id, d))[["x"]],
    `OLS predicted`     = predicted(shift),
    `OLS actual`        = coef(lm(y ~ x, d))[["x"]])
}

round(t(sapply(c(`shift 0` = 0, `shift 10` = 10, `shift 100` = 100), park)), 3)
```

### The same thing as a picture

The algebra says the between term takes over as the groups separate.
That is visible directly.  Left, the two groups share a centre; right,
the steady group is parked ten units away.  Dotted line is the mean of
$x$, dashed line joins the two group centres, solid line is the fit.

```{r}
#| fig-width: 9
#| fig-height: 4.2
set.seed(5)
m <- 200
draw <- function(shift, main) {
  xw <- rnorm(m, 0, 3);       yw <- 2 * xw + rnorm(m)      # wobbly: true slope 2
  xs <- rnorm(m, shift, 0.2); ys <- rnorm(m)               # steady: true slope 0
  x <- c(xw, xs); y <- c(yw, ys)
  plot(x, y, pch = 16, cex = .5, col = c(rep("#9E431855", m), rep("#0E6473", m)),
       xlab = "x", ylab = "y", main = main)
  abline(v = mean(x), lty = 3, col = "grey50")
  segments(mean(xw), mean(yw), mean(xs), mean(ys), lty = 2, lwd = 2, col = "grey30")
  points(c(mean(xw), mean(xs)), c(mean(yw), mean(ys)), pch = 21, bg = "white", cex = 1.4)
  abline(lm(y ~ x), lwd = 2, col = "#0E6473")
  dmu <- mean(xs) - mean(xw)
  Vb  <- 0.25 * dmu^2                       # between variance
  Vw  <- 0.5 * (var(xs) + var(xw))          # within variance
  c(`slope inside wobbly` = unname(coef(lm(yw ~ xw))[2]),
    `slope inside steady` = unname(coef(lm(ys ~ xs))[2]),
    `between slope`       = unname((mean(ys) - mean(yw)) / dmu),
    `weight on between %` = 100 * Vb / (Vb + Vw),
    `OLS`                 = unname(coef(lm(y ~ x))[2]))
}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
a <- draw(0,  "groups share a centre")
b <- draw(10, "groups far apart in x")
par(op)
round(cbind(`share a centre` = a, `far apart` = b), 3)
```

Read the two columns.  The slope inside the wobbly cloud is about 2 in
both, and the slope inside the steady strip is indistinguishable from
zero in both --- $0.007$ and $-0.288$, the second noisy because a
regression on an $x$ that barely varies has a standard error of roughly
$0.35$.  The groups themselves are unchanged.  What changes is the
weight on the between comparison: 0.03% when the centres coincide, 85%
when they are ten apart.  So the fit lands on 2.01 in the first panel
and 0.29 in the second.

Notice the between slope reported for the left panel, 0.882.  It is
meaningless there, and harmlessly so: with the centres nearly on top of
each other $\Delta\mu\approx 0$, the ratio $\Delta\bar y/\Delta\mu$
divides by almost nothing, and its weight $\tfrac14\Delta\mu^2$ is
almost nothing too.  An unstable quantity multiplied by a vanishing
weight contributes nothing.

The picture also shows why the strip has influence in the right panel
despite barely moving.  Influence is distance from the mean of $x$, the
dotted line, and the strip sits ten units from it.  What it contributes
is not its own slope but its position: it anchors one end of the dashed
line.  A group dummy removes that channel and returns the within answer.

Predicted and actual agree in every row.  Fixed effects is flat at 1.984
against a theoretical 1.991, the gap being sampling noise.  Plain OLS
runs from 1.983 to 0.004 on identical true slopes, partly because the
steady half gains weight as it leaves the grand mean, and partly because
a second channel opens that is not weighting at all: with the halves at
different levels of $x$, part of the fitted line is a comparison
*between* them.  So "the weight is how much $x$ moves" describes the
demeaned case; it is exact once unit effects are removed and false in
general.

### When it does not matter

Weighting only distorts when the effect is related to the weight.  Give
every unit its own slope, let units differ in how much $x$ moves, and
compare two designs: in (A) the slopes are unrelated to that movement,
in (B) the slope is larger where $x$ moves more.  Both average near one.
Design (A) is the $s_i$-versus-$b_i$ story from earlier, run again as a
panel instead of a cross-section: true slopes drawn independently of the
weight, so the weighted and unweighted averages come out close together.

```{r}
set.seed(7)
sdX <- runif(N, 0.2, 3)          # how much x moves, unit by unit

panel <- function(b) {           # build a panel whose unit slopes are b
  h   <- data.frame(id = rep(1:N, each = T),
                    x  = rnorm(N * T, 0, rep(sdX, each = T)))
  h$y <- rep(b, each = T) * h$x + rep(rnorm(N), each = T) + rnorm(N * T)
  h
}
compare <- function(h, b) {
  w <- tapply(resid(feols(x ~ 1 | id, h))^2, h$id, sum)   # the weights
  c(`FE estimate`            = coef(feols(y ~ x | id, h))[["x"]],
    `true slopes, plain avg` = mean(b),
    `true slopes, weighted`  = sum(w * b) / sum(w))
}

b1 <- rnorm(N, 1, 1)             # (A) slopes unrelated to movement
b2 <- 1 + 0.8 * (sdX - 1.6)      # (B) slopes larger where x moves more
h1 <- panel(b1); h2 <- panel(b2)

round(cbind(`(A) unrelated` = compare(h1, b1),
            `(B) aligned`   = compare(h2, b2)), 4)
```

In (A) all three numbers agree to within sampling noise: the weighting
is there and costs nothing.  In (B) the estimate is 1.53 while the plain
average of the true slopes is 1.01, and the weighted average, 1.54, is
what the estimator was tracking all along.  None of this is a defect ---
a unit whose $x$ barely moves genuinely carries little information about
how $y$ responds, and leaning on the units that do move is efficient.

Design (B) is the continuous version of what Słoczyński describes for a
binary treatment, worked through earlier in this chapter: the same
weighting, aggregated to two groups instead of spread over a continuum.

With homogeneous effects the weighting is harmless.  With heterogeneous
effects the question is never whether OLS is weighting, only whether we
know what it weighted.
