# Panel Data Models

```{r}
#| include: false
library(fixest)
```

## Background

Panel data are repeated observations on the same "unit" (we call that
cross-sectional units) over time.  Panel data models are used since it
has advantages over cross-sectional models or time-series models.  The
major advantage it has over cross-sectional models is: **Controlling for individual heterogeneity**.

For example, Baltagi and Levin (1992) studies cigarette demand across
46 states for the years 1963-88.  Consumption is modeled as a function
of lagged consumption, price and income.  These variables vary across
states and time.  However, there are other variables that may be
time-invariant ($z_i$) or individual-invariant ($w_t$).  Examples of $z_i$
are religion and, in a panel of adults, completed education.  For example, Utah, as a Mormon state, has
very low cigarette demand due to religious reason.  Generally we
consider it does not change over time or change very little over time.
Examples of $w_t$ include cigarette commercials on national TV or
radio.  Panel data models are able to control for individual
heterogeneity (or cross-sectional heterogeneity) while cross-sectional
models are not.  In many (or most) social science studies, there is
"unobserved effects" which is embedded in the error term. In other
words, we have "omitted variable" problem.  Without controlling for
it, the estimation results are generally biased and inconsistent.

Panel data give more informative data, more variability and less
collinearity among variables.  Often time series data suffers from
multicollinearity, while panel data has more variability from its
cross-sectional units.

The basic unobserved effects model can be written as:
$$
y_{it}={\bf {x_{it} \beta}} + c_i + u_{it}
$$

where $i$ indexes "units" (can be people, firms, or households,
etc.), and $t$ indexes time periods.

Comparing to regular cross-sectional models, there is an extra term
$c_i$.  $c_i$ can be treated as a random effect or fixed effect.
Traditionally it is distinguished by whether it is estimated as a
random variable or a parameter.  Modern econometrics tends to
distinguish by the correlation between $c_i$ and ${\bf {x_{it}}}$.
If there is no correlation between $c_i$ and ${\bf {x_{it}}}$,
then it's a random effect; otherwise, it's a fixed effect.

It is true that in some cases it is hard to justify that there is
no correlation between $c_i$ and ${\bf {x_{it}}}$, which is one reason
that people in economics tend to use fixed-effect models.  However,
fixed-effect models have its own limitations.  One of them is: It is
hard to justify that ALL the cross-sectional units' characteristics
other than those already in the model do not change over time.

## Random Effect Methods

A random effect model puts $c_i$ into the error term, then estimate by
FGLS (feasible GLS).  Random effect model is sometimes called
"Error-components model" since the overall error term is divided into
an individual level error term ($c_i$) and individual-time level error
term ($u_{it}$).


Under a random effect model, the variance-covariance matrix of the
error term becomes:


$$
\bf \Omega=
\begin{bmatrix} \sigma_{c}^2 + \sigma_u^2 & \sigma_{c}^2 & \cdots & \sigma_{c}^2
  \\ \sigma_{c}^2 & \sigma_{c}^2 + \sigma_u^2 & \cdots & \sigma_{c}^2 \\ \vdots &
  \vdots & \ddots & \vdots \\ \sigma_{c}^2  & \sigma_{c}^2  & \cdots & \sigma_{c}^2 + \sigma_u^2
\end{bmatrix}
$$


The random effect estimator is:

$$ \bf \hat \beta_{RE} = (\sum_{i=1}^N X_i' \hat \Omega^{-1} X_i)^{-1} (\sum_{i=1}^N X_i' \hat \Omega^{-1} y_i)$$

It is a special case of GLS.  Before calculating $\hat \beta_{RE}$, $
\sigma_{c}^2$ and $ \sigma_{u}^2$ need to be estimated.  This is
generally done by using residuals from within and between regressions
(the Swamy-Arora method, the default in most software), or from pooled
OLS.

## Fixed Effect Methods

Random effect methods assume that $c_i$ be orthogonal to $\bf x_{it}$.
In many applications, the whole point of using panel data is to allow
$c_i$ be correlated with $\bf x_{it}$.  We need fixed-effect models in
those cases.

For the same model as in the equation above, the fixed-effect
methods try to eliminate $c_i$ using fixed effects transformation, or
"within transformation".  The FE transformation is to "de-mean"
each observation by subtracting the group mean.  Basically subtracting
the group-mean equation below from the unit-level equation above:

$$
\bar y_i={\bf \bar x_i \beta} + c_i + \bar u_i
$$

where $\bar y_i $, etc., means group means.  For example, if we have 50 states
with state level data of cigarette consumption across years, then
groups means states, and we have 50 group means which are state level
average cigarette consumption across years.

The reason for "demeaning" is to remove $c_i$ from final estimation.
The "demeaned" regression equation is the final regression used in
estimation:

$$
(y_{it}-\bar y_i) = ({\bf x}_{it}-{\bf \bar x}_i)\beta + (u_{it}-\bar u_i).
$$

The unit-specific effect $c_i$ cancels because it is constant over $t$
($c_i - \bar c_i = 0$). Writing the within-transformed variables as
$\ddot y_{it}=y_{it}-\bar y_i$ and $\ddot{\bf x}_{it}={\bf x}_{it}-{\bf \bar x}_i$,
OLS on the demeaned data gives the **fixed-effects (within) estimator**
(here ${\bf x}_{it}$ is a $1\times k$ *row* vector, so
$\ddot{\bf x}_{it}'\ddot{\bf x}_{it}$ is the $k\times k$ outer product)

$$
\hat\beta_{FE} = \left(\sum_i \sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it}\right)^{-1}
\sum_i \sum_t \ddot{\bf x}_{it}'\ddot y_{it}.
$$

This is consistent for $\beta$ even when $c_i$ is correlated with
${\bf x}_{it}$, which is the key advantage of FE over random effects.
The price is that any time-invariant regressor (the $z_i$ above) is also
swept out by the demeaning and cannot be estimated. The usual variance
estimator is $\hat V(\hat\beta_{FE})=\hat\sigma_u^2(\sum_i\sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it})^{-1}$
with $\hat\sigma_u^2$ using degrees of freedom $NT-N-k$; in practice one
clusters by unit to allow serial correlation within $i$.

## Fixed effects or random effects: the Hausman test

Random effects is more efficient (it uses both within and between
variation) but is consistent only if $c_i$ is uncorrelated with
${\bf x}_{it}$; fixed effects is consistent either way. The Hausman test
compares the two estimators:

$$
H = (\hat\beta_{FE}-\hat\beta_{RE})'\,[\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})]^{-1}\,(\hat\beta_{FE}-\hat\beta_{RE}) \;\sim\; \chi^2_k
$$

under the null that random effects is consistent. A large $H$ (small
$p$-value) favors fixed effects. In R this is `phtest(fe, re)` from the
`plm` package; in Stata, `hausman fe re`.

**Caveat: this classic form requires RE to be efficient.** The formula above relies on $\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})$ being positive semi-definite, which in turn requires the RE estimator to actually be the *efficient* estimator under the null -- true only under the RE model's homoscedasticity and no-serial-correlation assumptions. If you are using robust or cluster-robust covariance matrices (the usual practice when errors are heteroscedastic or serially correlated), RE is generally no longer efficient, the covariance difference can fail to be positive definite, and the classic Hausman test statistic can be negative or otherwise ill-behaved. In that case, use a regression-based Hausman test instead: augment the RE (or pooled) regression with the within-unit means of the time-varying regressors (Mundlak's approach) or with the demeaned regressors themselves (Wooldridge's regression-based test), and test the added coefficients jointly with a standard (robust) Wald test -- this is valid under heteroscedasticity/clustering and reduces to the same idea as the classic Hausman test under its assumptions.

## Collapse to unit means, or keep the long format?

A practical question comes up whenever we hold a panel.  Should we
average each unit over time and run $N$ regressions, or keep all $NT$
rows and cluster the standard errors by unit?  Many people treat this
as a trade-off, and assume the long format must carry more
information.  It usually does not.  The answer turns entirely on where
the regressors live.

Take the case where every regressor is time-invariant.  The model is

$$
y_{it}={\bf z}_i \gamma + c_i + u_{it}
$$

and we have two options: pooled OLS on the $NT$ rows, or OLS on the
$N$ unit means $\bar y_i$.  With a balanced panel the two give the same
$\hat\gamma$.  Not close -- identical.  Random effects returns the same
number as well.  Wooldridge (2010) states the reason plainly: because
$c_i$ is there, new observations within a unit carry no information
about $\gamma$ beyond how they move the unit average $\bar y_i$.  In
effect we only have $N$ useful pieces of information.

In other words, averaging discards nothing, because ${\bf z}_i$ is
already constant over $t$ and the within variation was never available
to $\gamma$ in the first place.

What does differ is the degrees of freedom.  The unit-mean regression
uses $N-k$.  Pooled OLS reports $NT-k$, and that is the mistake.
Moulton (1990) is the classic illustration, and Donald and Lang (2007)
show that the $t_{N-k}$ reference distribution is the one that keeps
the size right.  Clustering by unit repairs the standard error when $N$
is large, but it does not repair the degrees of freedom when $N$ is
small.

Now put a time-varying ${\bf x}_{it}$ into the model.  The two options
stop agreeing, and not only for the coefficient on ${\bf x}_{it}$.
Every coefficient moves, the one on ${\bf z}_i$ included.

The reason is the between-within split from the sections above.
Averaging to unit means throws away all within variation, so what
survives is the between estimator.  Pooled OLS keeps both sources and
forces a single coefficient on them, so it returns a variance-weighted
average of the within and the between relationship.  The two agree only
when the within and between coefficients are equal.  That is precisely
the null the Hausman test examines.  So this is not a new problem.  It
is the FE-versus-RE question in different clothes.

| Situation | Collapsed vs long |
|---|---|
| All regressors time-invariant, balanced panel | Same coefficients; only the degrees of freedom differ |
| Any time-varying regressor | All coefficients differ, including those on ${\bf z}_i$ |
| Unbalanced panel | Differ regardless, since pooled OLS weights each unit by $T_i$ |

The reconciliation is Mundlak's device, the same one used just above
for the regression-based Hausman test.  Add the unit means
${\bf \bar x}_i$ to the long regression alongside ${\bf x}_{it}$.  The
within and between variation then carry separate coefficients, and the
coefficient on ${\bf \bar x}_i$ reproduces what the collapsed
regression estimates.  We keep the within variation and we recover the
between answer, in one regression.

The useful way to remember it: collapse freely when every regressor is
constant within the group, and never when one is not.

## What the long format buys

The section above is mostly a warning.  It says when collapsing is
harmless and when it is not.  The other side is worth seeing directly,
so we simulate a panel where we know the truth.  We use `fixest` here;
in Stata the same three regressions are `regress`, `xtreg, fe` and
`xtreg, re`.

### First, the equivalence

Take 1000 units over 6 periods, one time-invariant regressor $z_i$, and
a true $\gamma=2$.

```{r}
set.seed(42)
N <- 1000; T <- 6

z   <- rnorm(N)
bal <- data.frame(id = rep(1:N, each = T), t = rep(1:T, N),
                  z  = rep(z, each = T), c = rep(rnorm(N), each = T))
bal$y <- 2 * bal$z + bal$c + rnorm(N * T)
agg   <- aggregate(cbind(y, z) ~ id, data = bal, FUN = mean)

print(c(long      = coef(feols(y ~ z, bal))[["z"]],
        collapsed = coef(feols(y ~ z, agg))[["z"]]), digits = 12)
```

The two agree to every digit printed.  That is the result quoted above,
not an accident of the seed.

Now break the balance.  We drop the later periods for units with high
$z_i$, so $T_i$ varies, and varies with the regressor.

```{r}
unb  <- bal[!(bal$z > 0.5 & bal$t > 2), ]
agg2 <- aggregate(cbind(y, z) ~ id, data = unb, FUN = mean)

print(c(long      = coef(feols(y ~ z, unb))[["z"]],
        collapsed = coef(feols(y ~ z, agg2))[["z"]]), digits = 12)
```

They no longer agree.  Pooled OLS weights each unit by $T_i$; the
collapsed regression counts each unit once.  Neither is wrong.  They
answer different questions, and we should know which one we asked.

### Then, what collapsing costs

Now the case that matters.  We let $x_{it}$ vary over time, and we make
$c_i$ correlated with $\bar x_i$.  That is the situation fixed effects
exists for.  The truth is $\beta=1$.

```{r}
mu <- rnorm(N)
d  <- data.frame(id = rep(1:N, each = T),
                 x  = rep(mu, each = T) + rnorm(N * T),
                 c  = rep(mu + rnorm(N), each = T))
d$y    <- 1 * d$x + d$c + rnorm(N * T)
d$xbar <- ave(d$x, d$id)
dagg   <- aggregate(cbind(y, x) ~ id, data = d, FUN = mean)

between <- feols(y ~ x,        data = dagg)
pooled  <- feols(y ~ x,        data = d, cluster = ~id)
within  <- feols(y ~ x | id,   data = d)
mundlak <- feols(y ~ x + xbar, data = d, cluster = ~id)

pull <- function(m, v) c(coef(m)[[v]], se(m)[[v]])
tab <- rbind(`between (collapsed)` = pull(between, "x"),
             `pooled OLS`          = pull(pooled,  "x"),
             `within / FE`         = pull(within,  "x"),
             `Mundlak: x`          = pull(mundlak, "x"),
             `Mundlak: xbar`       = pull(mundlak, "xbar"))
colnames(tab) <- c("estimate", "std. error")
round(tab, 3)
```

Read the table down.  The collapsed regression returns about 1.85,
badly biased, because it is the between estimator and $\bar x_i$
carries $c_i$ with it.  Pooled OLS returns about 1.48 -- better, since
some within variation is mixed in, but still inconsistent.  The within
estimator returns about 0.99.  Only the long format can compute it.
Once we collapse there is nothing left to demean.

Mundlak's regression delivers both at once.  The coefficient on
$x_{it}$ reproduces the within estimate to three decimals.  The
coefficient on $\bar x_i$ is about 0.85, which is the between estimate
minus the within estimate.  In words, that second coefficient *is* the
Hausman contrast, and its $t$-statistic is the Hausman test.  A
collapsed regression cannot produce it, because it never sees the two
sources of variation separately.

So the argument for the long format is not precision.  It is that
demeaning, the Hausman comparison, unit-specific slopes, lagged
dependent variables and anything else built on within variation simply
do not exist once the panel is collapsed.  Dynamic models are the
extreme case, and they get their own chapter.
