# Panel Data Models

```{r}
#| include: false
library(fixest)
```

## Background

Panel data are repeated observations on the same "unit" (we call that
cross-sectional units) over time.  Panel data models are used since it
has advantages over cross-sectional models or time-series models.  The
major advantage it has over cross-sectional models is: **Controlling for individual heterogeneity**.

For example, Baltagi and Levin (1992) studies cigarette demand across
46 states for the years 1963-88.  Consumption is modeled as a function
of lagged consumption, price and income.  These variables vary across
states and time.  However, there are other variables that may be
time-invariant ($z_i$) or individual-invariant ($w_t$).  Examples of $z_i$
are religion and, in a panel of adults, completed education.  For example, Utah, as a Mormon state, has
very low cigarette demand due to religious reason.  Generally we
consider it does not change over time or change very little over time.
Examples of $w_t$ include cigarette commercials on national TV or
radio.  Panel data models are able to control for individual
heterogeneity (or cross-sectional heterogeneity) while cross-sectional
models are not.  In many (or most) social science studies, there is
"unobserved effects" which is embedded in the error term. In other
words, we have "omitted variable" problem.  Without controlling for
it, the estimation results are generally biased and inconsistent.

Panel data give more informative data, more variability and less
collinearity among variables.  Often time series data suffers from
multicollinearity, while panel data has more variability from its
cross-sectional units.

The basic unobserved effects model can be written as:
$$
y_{it}={\bf {x_{it} \beta}} + c_i + u_{it}
$$

where $i$ indexes "units" (can be people, firms, or households,
etc.), and $t$ indexes time periods.

Three kinds of regressor appear in this chapter, and it pays to name
them once.  ${\bf x}_{it}$ varies across units *and* time --- price,
income, and in most applications the treatment itself.  $z_i$ varies
across units only, so within a unit it is fixed over time; religion and
completed education above are the examples.  $w_t$ varies across time
only, so in any period it is common to every unit; the national TV
campaign is one.  Which group a variable falls into is what decides
almost everything that follows, including which coefficients fixed
effects can estimate and how OLS weights the data.

Instruments do not appear until the panel turns dynamic.  When they do,
they are written ${\bf Z}$, which is a different object from the
time-invariant $z_i$ here --- see the
[Endogeneity](endogeneity.qmd), [GMM](gmm.qmd) and
[Dynamic Panel Data](dynamic-panel.qmd) chapters.

Comparing to regular cross-sectional models, there is an extra term
$c_i$.  $c_i$ can be treated as a random effect or fixed effect.
Traditionally it is distinguished by whether it is estimated as a
random variable or a parameter.  Modern econometrics tends to
distinguish by the correlation between $c_i$ and ${\bf {x_{it}}}$.
If there is no correlation between $c_i$ and ${\bf {x_{it}}}$,
then it's a random effect; otherwise, it's a fixed effect.

It is true that in some cases it is hard to justify that there is
no correlation between $c_i$ and ${\bf {x_{it}}}$, which is one reason
that people in economics tend to use fixed-effect models.  However,
fixed-effect models have its own limitations.  One of them is: It is
hard to justify that ALL the cross-sectional units' characteristics
other than those already in the model do not change over time.

## Random Effect Methods

A random effect model puts $c_i$ into the error term, then estimate by
FGLS (feasible GLS).  Random effect model is sometimes called
"Error-components model" since the overall error term is divided into
an individual level error term ($c_i$) and individual-time level error
term ($u_{it}$).


Under a random effect model, the variance-covariance matrix of the
error term becomes:


$$
\bf \Omega=
\begin{bmatrix} \sigma_{c}^2 + \sigma_u^2 & \sigma_{c}^2 & \cdots & \sigma_{c}^2
  \\ \sigma_{c}^2 & \sigma_{c}^2 + \sigma_u^2 & \cdots & \sigma_{c}^2 \\ \vdots &
  \vdots & \ddots & \vdots \\ \sigma_{c}^2  & \sigma_{c}^2  & \cdots & \sigma_{c}^2 + \sigma_u^2
\end{bmatrix}
$$


The random effect estimator is:

$$ \bf \hat \beta_{RE} = (\sum_{i=1}^N X_i' \hat \Omega^{-1} X_i)^{-1} (\sum_{i=1}^N X_i' \hat \Omega^{-1} y_i)$$

It is a special case of GLS.  Before calculating $\hat \beta_{RE}$, $
\sigma_{c}^2$ and $ \sigma_{u}^2$ need to be estimated.  This is
generally done by using residuals from within and between regressions
(the Swamy-Arora method, the default in most software), or from pooled
OLS.

## Fixed Effect Methods

Random effect methods assume that $c_i$ be orthogonal to $\bf x_{it}$.
In many applications, the whole point of using panel data is to allow
$c_i$ be correlated with $\bf x_{it}$.  We need fixed-effect models in
those cases.

For the same model as in the equation above, the fixed-effect
methods try to eliminate $c_i$ using fixed effects transformation, or
"within transformation".  The FE transformation is to "de-mean"
each observation by subtracting the group mean.  Basically subtracting
the group-mean equation below from the unit-level equation above:

$$
\bar y_i={\bf \bar x_i \beta} + c_i + \bar u_i
$$

where $\bar y_i $, etc., means group means.  For example, if we have 50 states
with state level data of cigarette consumption across years, then
groups means states, and we have 50 group means which are state level
average cigarette consumption across years.

The reason for "demeaning" is to remove $c_i$ from final estimation.
The "demeaned" regression equation is the final regression used in
estimation:

$$
(y_{it}-\bar y_i) = ({\bf x}_{it}-{\bf \bar x}_i)\beta + (u_{it}-\bar u_i).
$$

The unit-specific effect $c_i$ cancels because it is constant over $t$
($c_i - \bar c_i = 0$). Writing the within-transformed variables as
$\ddot y_{it}=y_{it}-\bar y_i$ and $\ddot{\bf x}_{it}={\bf x}_{it}-{\bf \bar x}_i$,
OLS on the demeaned data gives the **fixed-effects (within) estimator**
(here ${\bf x}_{it}$ is a $1\times k$ *row* vector, so
$\ddot{\bf x}_{it}'\ddot{\bf x}_{it}$ is the $k\times k$ outer product)

$$
\hat\beta_{FE} = \left(\sum_i \sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it}\right)^{-1}
\sum_i \sum_t \ddot{\bf x}_{it}'\ddot y_{it}.
$$

This is consistent for $\beta$ even when $c_i$ is correlated with
${\bf x}_{it}$, which is the key advantage of FE over random effects.
The price is that any time-invariant regressor (the $z_i$ above) is also
swept out by the demeaning and cannot be estimated. The usual variance
estimator is $\hat V(\hat\beta_{FE})=\hat\sigma_u^2(\sum_i\sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it})^{-1}$
with $\hat\sigma_u^2$ using degrees of freedom $NT-N-k$; in practice one
clusters by unit to allow serial correlation within $i$.

## Fixed effects or random effects: the Hausman test

Random effects is more efficient (it uses both within and between
variation) but is consistent only if $c_i$ is uncorrelated with
${\bf x}_{it}$; fixed effects is consistent either way. The Hausman test
compares the two estimators:

$$
H = (\hat\beta_{FE}-\hat\beta_{RE})'\,[\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})]^{-1}\,(\hat\beta_{FE}-\hat\beta_{RE}) \;\sim\; \chi^2_k
$$

under the null that random effects is consistent. A large $H$ (small
$p$-value) favors fixed effects. In R this is `phtest(fe, re)` from the
`plm` package; in Stata, `hausman fe re`.

**Caveat: this classic form requires RE to be efficient.** The formula above relies on $\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})$ being positive semi-definite, which in turn requires the RE estimator to actually be the *efficient* estimator under the null -- true only under the RE model's homoscedasticity and no-serial-correlation assumptions. If you are using robust or cluster-robust covariance matrices (the usual practice when errors are heteroscedastic or serially correlated), RE is generally no longer efficient, the covariance difference can fail to be positive definite, and the classic Hausman test statistic can be negative or otherwise ill-behaved. In that case, use a regression-based Hausman test instead: augment the RE (or pooled) regression with the within-unit means of the time-varying regressors (Mundlak's approach) or with the demeaned regressors themselves (Wooldridge's regression-based test), and test the added coefficients jointly with a standard (robust) Wald test -- this is valid under heteroscedasticity/clustering and reduces to the same idea as the classic Hausman test under its assumptions.

## Collapse to unit means, or keep the long format?

A practical question comes up whenever we hold a panel.  Should we
average each unit over time and run one regression on the $N$ resulting
rows, or keep all $NT$ rows and cluster the standard errors by unit?  Many people treat this
as a trade-off, and assume the long format must carry more
information.  It usually does not.  The answer turns entirely on where
the regressors live.

Take the case where every regressor is time-invariant.  The model is

$$
y_{it}={\bf z}_i \gamma + c_i + u_{it}
$$

and we have two options: pooled OLS on the $NT$ rows, or OLS on the
$N$ unit means $\bar y_i$.  With a balanced panel the two give the same
$\hat\gamma$.  Not close -- identical.  Random effects returns the same
number as well.  Wooldridge (2010) states the reason plainly: because
$c_i$ is there, new observations within a unit carry no information
about $\gamma$ beyond how they move the unit average $\bar y_i$.  In
effect we only have $N$ useful pieces of information.

In other words, averaging discards nothing, because ${\bf z}_i$ is
already constant over $t$ and the within variation was never available
to $\gamma$ in the first place.

What does differ is the degrees of freedom.  The unit-mean regression
uses $N-k$.  Pooled OLS reports $NT-k$, and that is the mistake.
Moulton (1990) is the classic illustration, and Donald and Lang (2007)
show that the $t_{N-k}$ reference distribution is the one that keeps
the size right.  Clustering by unit repairs the standard error when $N$
is large, but it does not repair the degrees of freedom when $N$ is
small.

Now put a time-varying ${\bf x}_{it}$ into the model.  The two options
stop agreeing, and not only for the coefficient on ${\bf x}_{it}$.
Every coefficient moves, the one on ${\bf z}_i$ included.

The reason is the between-within split from the sections above.
Averaging to unit means throws away all within variation, so what
survives is the between estimator.  Pooled OLS keeps both sources and
forces a single coefficient on them, so it returns a variance-weighted
average of the within and the between relationship.  The two agree only
when the within and between coefficients are equal.  That is precisely
the null the Hausman test examines.  So this is not a new problem.  It
is the FE-versus-RE question in different clothes.

| Situation | Collapsed vs long |
|---|---|
| All regressors time-invariant, balanced panel | Same coefficients; only the degrees of freedom differ |
| Any time-varying regressor | All coefficients differ, including those on ${\bf z}_i$ |
| Unbalanced panel | Differ regardless, since pooled OLS weights each unit by $T_i$ |

The reconciliation is Mundlak's device, the same one used just above
for the regression-based Hausman test.  Add the unit means
${\bf \bar x}_i$ to the long regression alongside ${\bf x}_{it}$.  The
within and between variation then carry separate coefficients.  The
coefficient on ${\bf x}_{it}$ is the within estimate, and the one on
${\bf \bar x}_i$ is the *difference* between the between and within
estimates --- so their sum is what the collapsed regression reports.  We
keep the within variation and recover the between answer, in one
regression.

The useful way to remember it: collapse freely when every regressor is
constant within the group, and never when one is not.

## What the long format buys

The section above is mostly a warning.  It says when collapsing is
harmless and when it is not.  The other side is worth seeing directly,
so we simulate a panel where we know the truth.  We use `fixest` here;
in Stata the same regressions are `regress`, `xtreg, be` and
`xtreg, fe`.

### First, the equivalence

Take 1000 units over 6 periods, one time-invariant regressor $z_i$, and
a true $\gamma=2$.

```{r}
set.seed(42)
N <- 1000; T <- 6

z   <- rnorm(N)
bal <- data.frame(id = rep(1:N, each = T), t = rep(1:T, N),
                  z  = rep(z, each = T), c = rep(rnorm(N), each = T))
bal$y <- 2 * bal$z + bal$c + rnorm(N * T)
agg   <- aggregate(cbind(y, z) ~ id, data = bal, FUN = mean)

print(c(long      = coef(feols(y ~ z, bal))[["z"]],
        collapsed = coef(feols(y ~ z, agg))[["z"]]), digits = 12)
```

The two agree to every digit printed.  That is the result quoted above,
not an accident of the seed.

Now break the balance.  We drop the later periods for units with high
$z_i$, so $T_i$ varies, and varies with the regressor.

```{r}
unb  <- bal[!(bal$z > 0.5 & bal$t > 2), ]
agg2 <- aggregate(cbind(y, z) ~ id, data = unb, FUN = mean)

print(c(long      = coef(feols(y ~ z, unb))[["z"]],
        collapsed = coef(feols(y ~ z, agg2))[["z"]]), digits = 12)
```

They no longer agree.  Pooled OLS weights each unit by $T_i$; the
collapsed regression counts each unit once.  Neither is wrong.  They
answer different questions, and we should know which one we asked.
The $T_i$ here is a special case of a general rule, which the last
section of this chapter sets out.

### Then, what collapsing costs

Now the case that matters.  We let $x_{it}$ vary over time, and we make
$c_i$ correlated with $\bar x_i$.  That is the situation fixed effects
exists for.  The truth is $\beta=1$.

```{r}
mu <- rnorm(N)
d  <- data.frame(id = rep(1:N, each = T),
                 x  = rep(mu, each = T) + rnorm(N * T),
                 c  = rep(mu + rnorm(N), each = T))
d$y    <- 1 * d$x + d$c + rnorm(N * T)
d$xbar <- ave(d$x, d$id)
dagg   <- aggregate(cbind(y, x) ~ id, data = d, FUN = mean)

between <- feols(y ~ x,        data = dagg)
pooled  <- feols(y ~ x,        data = d, cluster = ~id)
within  <- feols(y ~ x | id,   data = d)
mundlak <- feols(y ~ x + xbar, data = d, cluster = ~id)

pull <- function(m, v) c(coef(m)[[v]], se(m)[[v]])
tab <- rbind(`between (collapsed)` = pull(between, "x"),
             `pooled OLS`          = pull(pooled,  "x"),
             `within / FE`         = pull(within,  "x"),
             `Mundlak: x`          = pull(mundlak, "x"),
             `Mundlak: xbar`       = pull(mundlak, "xbar"))
colnames(tab) <- c("estimate", "std. error")
round(tab, 3)
```

Read the table down.  The collapsed regression returns about 1.85,
badly biased, because it is the between estimator and $\bar x_i$
carries $c_i$ with it.  Pooled OLS returns about 1.48 -- better, since
some within variation is mixed in, but still inconsistent.  The within
estimator returns about 0.99.  Only the long format can compute it.
Once we collapse there is nothing left to demean.

Mundlak's regression delivers both at once.  The coefficient on
$x_{it}$ reproduces the within estimate to three decimals.  The
coefficient on $\bar x_i$ is about 0.85, which is the between estimate
minus the within estimate.  In words, that second coefficient *is* the
Hausman contrast, and its $t$-statistic is the Hausman test.  A
collapsed regression cannot produce it, because it never sees the two
sources of variation separately.

So the argument for the long format is not precision.  It is that
demeaning, the Hausman comparison, unit-specific slopes, lagged
dependent variables and anything else built on within variation simply
do not exist once the panel is collapsed.  Dynamic models are the
extreme case, and they get their own chapter,
[Dynamic Panel Data](dynamic-panel.qmd).

## How OLS weights the data

Twice above we said OLS weights units unequally.  By $T_i$ when the
panel is unbalanced, and by variance when it mixes within and between
variation.  Neither is a panel-specific quirk.  Both follow from one
rule, and the rule is worth stating on its own.

Let $V$ be the regressor we care about, and let $\tilde V$ be $V$ after
residualising on every other regressor in the model.  By the
Frisch--Waugh--Lovell theorem,

$$
\hat\beta = \frac{\sum_i \sum_t \tilde V_{it}\, y_{it}}
                 {\sum_i \sum_t \tilde V_{it}^{2}} .
$$

In words: an observation enters the numerator weighted by
$\tilde V_{it}$, to the *first* power, and signed.  Note what
$\tilde V_{it}$ is and is not.  A single observation has no variation of
its own --- variation is a property of a collection.  What it has is a
**deviation**: how far its regressor sits from what the rest of the
model predicts for it.  With only an intercept in the model that is
distance from the grand mean; with unit effects it is distance from the
unit's own mean.  Which mean is being subtracted matters enormously, and
we return to it below.  These weights sum to zero, and an observation
whose regressor is fully explained by the other regressors carries no
weight at all.

The squares come from asking a different question.  Weights on
*outcomes* are one object; weights on *effects* are another.  Let the
effect vary by unit, so $y_{it}=\beta_i V_{it}+\ldots$ with the
remainder orthogonal to $\tilde V$.  Substituting into the numerator
gives $\sum_i \beta_i \sum_t \tilde V_{it}V_{it}$.  Over the whole
sample $\sum_i\sum_t \tilde V_{it}V_{it}=\sum_i\sum_t \tilde V_{it}^2$,
because $\tilde V$ is orthogonal to the fitted part of $V$.

Whether that also holds *unit by unit* is a separate question, and it is
the one that licenses reading $\sum_t \tilde V_{it}^2$ as unit $i$'s
weight.  Demeaning delivers it by direct expansion,

$$
\sum_t (V_{it}-\bar V_i)\,V_{it}
= \sum_t V_{it}^2 - T\bar V_i^{\,2}
= \sum_t (V_{it}-\bar V_i)^2 ,
$$

so with unit effects in the model, and only then,

$$
\hat\beta = \frac{\sum_i \beta_i \sum_t \tilde V_{it}^2}
                 {\sum_i \sum_t \tilde V_{it}^2},
$$

a weighted average of the unit-level effects.  The square is not assumed
anywhere.  It is what a first-power weight on the outcome becomes once
the outcome is rewritten in terms of an effect.

Under a coarser control the sample total still matches but the per-unit
equality fails, and the tidy reading breaks down.  Hazlett and Shinkre
make this the centre of their treatment; the Angrist row in the table
below assumes the saturated case.

Both objects are developed at length in
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html).
The distinction earns its keep there: signed outcome weights can turn
negative when a linear probability model pushes $\hat d(x)$ outside
$[0,1]$, and squared effect weights, being non-negative, cannot express
that at all.

That sum is the whole story, and in general it is as far as we can go.
$\tilde V_{it}$ carries both subscripts even when $V$ itself does not,
so $\sum_t \tilde V_{it}^2$ usually will not factor into anything
tidier.  What changes from model to model is whether it happens to
simplify.

| Model | $\sum_t \tilde V_{it}^2$ simplifies to |
|---|---|
| $z_i$ alone, nothing else varying within the unit | $T_i\,\tilde z_i^2$ |
| a time-varying $x_{it}$, unit effects removed | $(T-1)\,\widehat{\text{Var}}_i(x)$ |
| binary $d$ with saturated controls | $n_x\,\hat d(x)\,(1-\hat d(x))$ |

Now add a time-varying $x_{it}$ alongside $z_i$ in the first model, and
the simplification is gone.  Residualising $z_i$ on something that moves
within the unit makes $\tilde z$ move within the unit as well, so we are
back to $\sum_t \tilde z_{it}^2$ with nothing to factor out.  This is
worth stating because the $T_i$ form from the previous section looks
like the general case and is not.  It is the one-regressor special case.

The last row is Angrist ([1998](https://doi.org/10.2307/2998558)).
Strata are weighted by $\hat d(x)(1-\hat d(x))$, largest where treatment
is closest to a coin flip.  Same rule, different regressor.

### When the weighting changes the answer

The weighting is always happening.  It changes the answer only when the
effect is heterogeneous *and* the heterogeneity lines up with the
weight.  The claim to be checked is this: a regression does not report
*the* effect, it reports a weighted average of the effects, and those
are two different numbers.

We cannot check that on real data, because the individual effects are
never observed.  So we plant them.  Give every unit $i$ its own slope
$\beta_i$, generate $y$ from those slopes, run one regression, and see
which average of the planted slopes the estimate lands on.  In the next chunk only one
estimator is run; anything else reported is not an estimate at all, but
the planted slopes averaged some way.  Fixed
effects is used because in a panel the natural place for heterogeneity
is across units, and it is still OLS --- OLS with the unit dummies
partialled out --- so the rule above applies to it unchanged.

Start with the extreme case, where there is nothing to disentangle.
Half the units barely move and have a true slope of zero.  The other
half move a great deal and have a true slope of two.  The average unit's
slope is therefore one.

```{r}
set.seed(1)
half <- N / 2
g <- data.frame(id   = rep(1:N, each = T),
                move = rep(c(rep(0.2, half), rep(3, half)), each = T),
                b    = rep(c(rep(0,   half), rep(2, half)), each = T))
g$x <- rnorm(N * T, 0, g$move)
g$y <- g$b * g$x + rep(rnorm(N), each = T) + rnorm(N * T)

wg <- tapply(resid(feols(x ~ 1 | id, g))^2, g$id, sum)
round(cbind(value = c(
  `average of the two true slopes`      = mean(c(0, 2)),
  `FE estimate`                         = coef(feols(y ~ x | id, g))[["x"]],
  `weight held by the steady half (%)`  = 100 * sum(wg[1:half]) / sum(wg))), 3)
```

The average slope is one and the regression reports 1.98.  The reason is
in the third line: the five hundred steady units hold under half of one
percent of the weight.

Be precise about which deviation that is.  The regression above carries
unit effects, so $\tilde x$ is the deviation from each unit's *own*
mean, and a steady unit's six observations sit almost on top of that
mean.  Squaring puts them nowhere in the sum, and the estimate becomes
very nearly a report on the wobbly half alone.  Nothing has gone wrong
in the arithmetic.  The units with a slope of zero were never really in
the regression.

Note that this is a statement about the *within-unit* deviation, not
about distance from the grand mean.  Plain OLS uses the latter, and
they are not the same quantity --- as the next subsection shows, they
happen to coincide here only because both halves were built around
zero.

Before going further, a warning about what this example does *not*
show.  The section is called "How OLS weights the data" and the
regression above is a fixed-effects regression.  It is tempting to
conclude that plain OLS would behave the same way.  It would not, and
the reason matters more than the example.

$\tilde V$ is a deviation from whatever mean the model removes.  With
unit effects in the model that is the unit's own mean, so the weight is
within-unit spread.  With only an intercept it is the grand mean.  Above,
both halves were built around zero, which quietly made the two readings
coincide.  Separate them and the algebra says exactly what happens.

Take two equally sized groups with means $\mu_g$, within-group variances
$\sigma_g^2$, slopes $\beta_g$, and group mean outcomes $\bar y_g$.  The
law of total covariance splits $\text{Cov}(x,y)$ and $\text{Var}(x)$
into a within part and a between part, giving

$$
\text{plim}\;\hat\beta_{OLS}
= \frac{\tfrac12\!\left(\beta_1\sigma_1^2+\beta_2\sigma_2^2\right)
        \;+\; \tfrac14\,\Delta\mu\,\Delta\bar y}
       {\tfrac12\!\left(\sigma_1^2+\sigma_2^2\right)
        \;+\; \tfrac14\,\Delta\mu^{2}} ,
$$

with $\Delta\mu=\mu_1-\mu_2$ and $\Delta\bar y=\bar y_1-\bar y_2$.  In
words, the coefficient is a variance-weighted average of two slopes:

$$
\beta_{within}=\frac{\beta_1\sigma_1^2+\beta_2\sigma_2^2}{\sigma_1^2+\sigma_2^2},
\qquad
\beta_{between}=\frac{\Delta\bar y}{\Delta\mu},
$$

the first weighted by the within variance $\tfrac12(\sigma_1^2+\sigma_2^2)$
and the second by the between variance $\tfrac14\Delta\mu^2$.

Everything now follows without running anything.  Fixed effects
estimates $\beta_{within}$, which here is
$(0\cdot 0.04 + 2\cdot 9)/9.04 = 1.991$ and does not involve $\mu_g$ at
all --- so parking a group anywhere cannot move it.  Plain OLS also
carries $\beta_{between}$, and in this design $\beta_{between}=0$: the
steady group's slope is zero and both intercepts are zero, so the two
group means sit at the same height and the line through them is flat.
As $\Delta\mu$ grows, the between weight $\tfrac14\Delta\mu^2$ swamps
the fixed within weight of $4.52$, and the OLS coefficient is dragged
from $1.991$ toward $0$ like $9/(4.52+\Delta\mu^2/4)$, the numerator
being $\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)=9$.

The simulation is a check on that, not the source of it.

```{r}
b_within <- (0 * 0.2^2 + 2 * 3^2) / (0.2^2 + 3^2)   # variance-weighted slope
V_within <- 0.5 * (0.2^2 + 3^2)                     # within variance
predicted <- function(shift)                        # beta_between = 0 here
  V_within * b_within / (V_within + 0.25 * shift^2)

park <- function(shift) {
  set.seed(1)
  x <- rnorm(N * T, 0, g$move) + rep(c(rep(shift, half), rep(0, half)), each = T)
  y <- g$b * x + rep(rnorm(N), each = T) + rnorm(N * T)
  d <- data.frame(id = g$id, x = x, y = y)
  c(`FE (theory 1.991)` = coef(feols(y ~ x | id, d))[["x"]],
    `OLS predicted`     = predicted(shift),
    `OLS actual`        = coef(lm(y ~ x, d))[["x"]])
}

round(t(sapply(c(`shift 0` = 0, `shift 10` = 10, `shift 100` = 100), park)), 3)
```

Predicted and actual agree in every row, and fixed effects is flat at
1.984 against a theoretical 1.991 --- the gap is sampling noise, not
structure.

Two lessons.  First, "the weight is how much $x$ moves" describes the
*demeaned* case; it is exact once unit effects are removed and false in
general.  Second, when groups sit at different levels of $x$, plain OLS
mixes in $\beta_{between}$, which answers a different question
altogether.  Only the within part is the subject of this section.

That is the mechanism in its starkest form.  A continuous version makes
the same point, and adds the control case the extreme example cannot
show --- what happens when the slope has nothing to do with the
movement.

Two designs.  In both, units differ in how much $x$ moves for them.  In
(A) the slopes are unrelated to that movement.  In (B) the slope is
larger for units whose $x$ moves more.  Both have an average slope near
one, so the designs differ *only* in whether the heterogeneity lines up
with the weight.

```{r}
set.seed(7)
sdX <- runif(N, 0.2, 3)          # how much x moves, unit by unit

panel <- function(b) {           # build a panel whose unit slopes are b
  h   <- data.frame(id = rep(1:N, each = T),
                    x  = rnorm(N * T, 0, rep(sdX, each = T)))
  h$y <- rep(b, each = T) * h$x + rep(rnorm(N), each = T) + rnorm(N * T)
  h
}
compare <- function(h, b) {
  w <- tapply(resid(feols(x ~ 1 | id, h))^2, h$id, sum)   # the weights
  c(`FE estimate`            = coef(feols(y ~ x | id, h))[["x"]],
    `true slopes, plain avg` = mean(b),
    `true slopes, weighted`  = sum(w * b) / sum(w))
}

b1 <- rnorm(N, 1, 1)             # (A) slopes unrelated to movement
b2 <- 1 + 0.8 * (sdX - 1.6)      # (B) slopes larger where x moves more
h1 <- panel(b1); h2 <- panel(b2)

round(cbind(`(A) unrelated` = compare(h1, b1),
            `(B) aligned`   = compare(h2, b2)), 4)
```

Read down each column.  In (A) all three numbers agree to within
sampling noise: the weighting happens, and it costs nothing.  In (B) the
estimate is 1.53 while the plain average of the true slopes is 1.01.
The estimate has not gone wrong --- it matches the weighted average,
1.54, which is what it was reporting all along.  We were reading it as
an average.

The same mechanism as the two-group case, spread over a continuum.
Squaring is what drives it: a unit whose $x$ moves fifteen times as much
as another counts for roughly two hundred times as much.

None of this is a defect.  A unit whose $x$ barely moves genuinely
carries little information about how $y$ responds to $x$, and leaning on
the units that do move is the efficient thing to do.  It becomes a
problem only when the slope itself is related to the movement, which is
what (B) imposes and (A) does not.

Słoczyński ([2022](https://doi.org/10.1162/rest_a_00953)) makes the
same point for a binary treatment: the OLS coefficient is a convex
combination of the effect on the treated and the effect on the
untreated ([discussed here](https://xiangao.github.io/blog_book/ols-ate.html)),
and the *smaller* group receives the *larger* weight.  Writing $\rho$
for the treated share, the coefficient is
$(1-\rho)\,\text{ATT} + \rho\,\text{ATU}$, so the weight on the effect
for the treated is $1-\rho$.

Read that carefully, because it runs against intuition.  In a sample
that is 10% treated the weight on the ATT is $0.9$, so the coefficient
is mostly reporting the effect on the *treated tenth*, not on the
untreated majority.  The ATE, by contrast, is
$\rho\,\text{ATT} + (1-\rho)\,\text{ATU}$ --- the same two numbers with
the weights the other way round.  The two can be far apart: in a
simulation with 8.8% treated, an ATT of 2.39 and an ATU of 0.87, the
ATE is 1.00 while OLS returns 2.26.

The practical reading: with homogeneous effects the weighting is
harmless and we can ignore it.  With heterogeneous effects the question
is never whether OLS is weighting, only whether we know what it
weighted.
