# Panel Data Models

```{r}
#| include: false
library(fixest)
```

## Background

Panel data are repeated observations on the same "unit" (we call that
cross-sectional units) over time.  Panel data models are used since it
has advantages over cross-sectional models or time-series models.  The
major advantage it has over cross-sectional models is: **Controlling for individual heterogeneity**.

For example, Baltagi and Levin (1992) studies cigarette demand across
46 states for the years 1963-88.  Consumption is modeled as a function
of lagged consumption, price and income.  These variables vary across
states and time.  However, there are other variables that may be
time-invariant ($z_i$) or individual-invariant ($w_t$).  Examples of $z_i$
are religion and, in a panel of adults, completed education.  For example, Utah, as a Mormon state, has
very low cigarette demand due to religious reason.  Generally we
consider it does not change over time or change very little over time.
Examples of $w_t$ include cigarette commercials on national TV or
radio.  Panel data models are able to control for individual
heterogeneity (or cross-sectional heterogeneity) while cross-sectional
models are not.  In many (or most) social science studies, there is
"unobserved effects" which is embedded in the error term. In other
words, we have "omitted variable" problem.  Without controlling for
it, the estimation results are generally biased and inconsistent.

Panel data give more informative data, more variability and less
collinearity among variables.  Often time series data suffers from
multicollinearity, while panel data has more variability from its
cross-sectional units.

The basic unobserved effects model can be written as:
$$
y_{it}={\bf {x_{it} \beta}} + c_i + u_{it}
$$

where $i$ indexes "units" (can be people, firms, or households,
etc.), and $t$ indexes time periods.

Three kinds of regressor appear in this chapter, and it pays to name
them once.  ${\bf x}_{it}$ varies across units *and* time --- price,
income, and in most applications the treatment itself.  $z_i$ varies
across units only, so within a unit it is fixed over time; religion and
completed education above are the examples.  $w_t$ varies across time
only, so in any period it is common to every unit; the national TV
campaign is one.  Which group a variable falls into is what decides
almost everything that follows, including which coefficients fixed
effects can estimate and how OLS weights the data.

Instruments do not appear until the panel turns dynamic.  When they do,
they are written ${\bf Z}$, which is a different object from the
time-invariant $z_i$ here --- see the
[Endogeneity](endogeneity.qmd), [GMM](gmm.qmd) and
[Dynamic Panel Data](dynamic-panel.qmd) chapters.

Comparing to regular cross-sectional models, there is an extra term
$c_i$.  $c_i$ can be treated as a random effect or fixed effect.
Traditionally it is distinguished by whether it is estimated as a
random variable or a parameter.  Modern econometrics tends to
distinguish by the correlation between $c_i$ and ${\bf {x_{it}}}$.
If there is no correlation between $c_i$ and ${\bf {x_{it}}}$,
then it's a random effect; otherwise, it's a fixed effect.

It is true that in some cases it is hard to justify that there is
no correlation between $c_i$ and ${\bf {x_{it}}}$, which is one reason
that people in economics tend to use fixed-effect models.  However,
fixed-effect models have its own limitations.  One of them is: It is
hard to justify that ALL the cross-sectional units' characteristics
other than those already in the model do not change over time.

## Random Effect Methods

A random effect model puts $c_i$ into the error term, then estimate by
FGLS (feasible GLS).  Random effect model is sometimes called
"Error-components model" since the overall error term is divided into
an individual level error term ($c_i$) and individual-time level error
term ($u_{it}$).


Under a random effect model, the variance-covariance matrix of the
error term becomes:


$$
\bf \Omega=
\begin{bmatrix} \sigma_{c}^2 + \sigma_u^2 & \sigma_{c}^2 & \cdots & \sigma_{c}^2
  \\ \sigma_{c}^2 & \sigma_{c}^2 + \sigma_u^2 & \cdots & \sigma_{c}^2 \\ \vdots &
  \vdots & \ddots & \vdots \\ \sigma_{c}^2  & \sigma_{c}^2  & \cdots & \sigma_{c}^2 + \sigma_u^2
\end{bmatrix}
$$


The random effect estimator is:

$$ \bf \hat \beta_{RE} = (\sum_{i=1}^N X_i' \hat \Omega^{-1} X_i)^{-1} (\sum_{i=1}^N X_i' \hat \Omega^{-1} y_i)$$

It is a special case of GLS.  Before calculating $\hat \beta_{RE}$, $
\sigma_{c}^2$ and $ \sigma_{u}^2$ need to be estimated.  This is
generally done by using residuals from within and between regressions
(the Swamy-Arora method, the default in most software), or from pooled
OLS.

## Fixed Effect Methods

Random effect methods assume that $c_i$ be orthogonal to $\bf x_{it}$.
In many applications, the whole point of using panel data is to allow
$c_i$ be correlated with $\bf x_{it}$.  We need fixed-effect models in
those cases.

For the same model as in the equation above, the fixed-effect
methods try to eliminate $c_i$ using fixed effects transformation, or
"within transformation".  The FE transformation is to "de-mean"
each observation by subtracting the group mean.  Basically subtracting
the group-mean equation below from the unit-level equation above:

$$
\bar y_i={\bf \bar x_i \beta} + c_i + \bar u_i
$$

where $\bar y_i $, etc., means group means.  For example, if we have 50 states
with state level data of cigarette consumption across years, then
groups means states, and we have 50 group means which are state level
average cigarette consumption across years.

The reason for "demeaning" is to remove $c_i$ from final estimation.
The "demeaned" regression equation is the final regression used in
estimation:

$$
(y_{it}-\bar y_i) = ({\bf x}_{it}-{\bf \bar x}_i)\beta + (u_{it}-\bar u_i).
$$

The unit-specific effect $c_i$ cancels because it is constant over $t$
($c_i - \bar c_i = 0$). Writing the within-transformed variables as
$\ddot y_{it}=y_{it}-\bar y_i$ and $\ddot{\bf x}_{it}={\bf x}_{it}-{\bf \bar x}_i$,
OLS on the demeaned data gives the **fixed-effects (within) estimator**
(here ${\bf x}_{it}$ is a $1\times k$ *row* vector, so
$\ddot{\bf x}_{it}'\ddot{\bf x}_{it}$ is the $k\times k$ outer product)

$$
\hat\beta_{FE} = \left(\sum_i \sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it}\right)^{-1}
\sum_i \sum_t \ddot{\bf x}_{it}'\ddot y_{it}.
$$

This is consistent for $\beta$ even when $c_i$ is correlated with
${\bf x}_{it}$, which is the key advantage of FE over random effects.
The price is that any time-invariant regressor (the $z_i$ above) is also
swept out by the demeaning and cannot be estimated. The usual variance
estimator is $\hat V(\hat\beta_{FE})=\hat\sigma_u^2(\sum_i\sum_t \ddot{\bf x}_{it}'\ddot{\bf x}_{it})^{-1}$
with $\hat\sigma_u^2$ using degrees of freedom $NT-N-k$; in practice one
clusters by unit to allow serial correlation within $i$.

## Fixed effects or random effects: the Hausman test

Random effects is more efficient (it uses both within and between
variation) but is consistent only if $c_i$ is uncorrelated with
${\bf x}_{it}$; fixed effects is consistent either way. The Hausman test
compares the two estimators:

$$
H = (\hat\beta_{FE}-\hat\beta_{RE})'\,[\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})]^{-1}\,(\hat\beta_{FE}-\hat\beta_{RE}) \;\sim\; \chi^2_k
$$

under the null that random effects is consistent. A large $H$ (small
$p$-value) favors fixed effects. In R this is `phtest(fe, re)` from the
`plm` package; in Stata, `hausman fe re`.

**Caveat: this classic form requires RE to be efficient.** The formula above relies on $\hat V(\hat\beta_{FE})-\hat V(\hat\beta_{RE})$ being positive semi-definite, which in turn requires the RE estimator to actually be the *efficient* estimator under the null -- true only under the RE model's homoscedasticity and no-serial-correlation assumptions. If you are using robust or cluster-robust covariance matrices (the usual practice when errors are heteroscedastic or serially correlated), RE is generally no longer efficient, the covariance difference can fail to be positive definite, and the classic Hausman test statistic can be negative or otherwise ill-behaved. In that case, use a regression-based Hausman test instead: augment the RE (or pooled) regression with the within-unit means of the time-varying regressors (Mundlak's approach) or with the demeaned regressors themselves (Wooldridge's regression-based test), and test the added coefficients jointly with a standard (robust) Wald test -- this is valid under heteroscedasticity/clustering and reduces to the same idea as the classic Hausman test under its assumptions.

## Collapse to unit means, or keep the long format?

A practical question comes up whenever we hold a panel.  Should we
average each unit over time and run one regression on the $N$ resulting
rows, or keep all $NT$ rows and cluster the standard errors by unit?
Many people treat this as a trade-off and assume the long format must
carry more information.  It usually does not.  The answer turns entirely
on where the regressors live.

Take the case where every regressor is time-invariant, so the model is
$y_{it}={\bf z}_i \gamma + c_i + u_{it}$.  With a balanced panel, pooled
OLS on the $NT$ rows and OLS on the $N$ unit means give the same
$\hat\gamma$.  Not close --- identical.  Random effects returns the same
number as well.  Wooldridge (2010) states the reason plainly: because
$c_i$ is there, new observations within a unit carry no information
about $\gamma$ beyond how they move the unit average $\bar y_i$.  In
effect we only have $N$ useful pieces of information.  Averaging
discards nothing, because ${\bf z}_i$ was already constant over $t$ and
the within variation was never available to $\gamma$ in the first place.

What does differ is the degrees of freedom.  The unit-mean regression
uses $N-k$; pooled OLS reports $NT-k$, and that is the mistake.  Moulton
(1990) is the classic illustration, and Donald and Lang (2007) show that
$t_{N-k}$ is the reference distribution that keeps the size right.
Clustering by unit repairs the standard error when $N$ is large, but it
does not repair the degrees of freedom when $N$ is small.

Now put a time-varying ${\bf x}_{it}$ into the model.  The two options
stop agreeing, and not only for the coefficient on ${\bf x}_{it}$ ---
every coefficient moves, the one on ${\bf z}_i$ included.  Averaging to
unit means throws away all within variation, so what survives is the
between estimator.  Pooled OLS keeps both sources and forces a single
coefficient on them, so it returns a variance-weighted average of the
within and the between relationship.  The two agree only when the within
and between coefficients are equal, which is precisely the null the
Hausman test examines.  So this is not a new problem.  It is the
FE-versus-RE question in different clothes.

| Situation | Collapsed vs long |
|---|---|
| All regressors time-invariant, balanced panel | Same coefficients; only the degrees of freedom differ |
| Any time-varying regressor | All coefficients differ, including those on ${\bf z}_i$ |
| Unbalanced panel | Differ regardless, since pooled OLS weights each unit by $T_i$ |

The reconciliation is Mundlak's device, the same one used above for the
regression-based Hausman test.  Add the unit means ${\bf \bar x}_i$ to
the long regression alongside ${\bf x}_{it}$, and the within and between
variation carry separate coefficients: ${\bf x}_{it}$ picks up the
within estimate, ${\bf \bar x}_i$ picks up the *difference* between the
between and within estimates, so the two coefficients sum to what the
collapsed regression reports.  That second coefficient is the Hausman
contrast, and its $t$-statistic is the Hausman test.  We keep the within
variation and recover the between answer in one regression.

The useful way to remember it: collapse freely when every regressor is
constant within the group, and never when one is not.

## What the long format buys

The section above is mostly a warning.  The other side is worth seeing
directly, so we simulate panels where we know the truth.  We use
`fixest` here; in Stata the same regressions are `regress`, `xtreg, be`
and `xtreg, fe`.

### The equivalence, and where it fails

Take 1000 units over 6 periods, one time-invariant regressor $z_i$, and
a true $\gamma=2$.

```{r}
set.seed(42)
N <- 1000; T <- 6

z   <- rnorm(N)
bal <- data.frame(id = rep(1:N, each = T), t = rep(1:T, N),
                  z  = rep(z, each = T), c = rep(rnorm(N), each = T))
bal$y <- 2 * bal$z + bal$c + rnorm(N * T)
agg   <- aggregate(cbind(y, z) ~ id, data = bal, FUN = mean)

print(c(long      = coef(feols(y ~ z, bal))[["z"]],
        collapsed = coef(feols(y ~ z, agg))[["z"]]), digits = 12)
```

The two agree to every digit printed.  Now break the balance by dropping
the later periods for units with high $z_i$, so $T_i$ varies, and varies
with the regressor.

```{r}
unb  <- bal[!(bal$z > 0.5 & bal$t > 2), ]
agg2 <- aggregate(cbind(y, z) ~ id, data = unb, FUN = mean)

print(c(long      = coef(feols(y ~ z, unb))[["z"]],
        collapsed = coef(feols(y ~ z, agg2))[["z"]]), digits = 12)
```

They no longer agree.  Pooled OLS weights each unit by $T_i$; the
collapsed regression counts each unit once.  Neither is wrong --- they
answer different questions, and we should know which one we asked.  That
$T_i$ is a special case of a general rule, set out in the next section.

### What collapsing costs

Now the case that matters.  Let $x_{it}$ vary over time and make $c_i$
correlated with $\bar x_i$, which is the situation fixed effects exists
for.  The truth is $\beta=1$.

```{r}
mu <- rnorm(N)
d  <- data.frame(id = rep(1:N, each = T),
                 x  = rep(mu, each = T) + rnorm(N * T),
                 c  = rep(mu + rnorm(N), each = T))
d$y    <- 1 * d$x + d$c + rnorm(N * T)
d$xbar <- ave(d$x, d$id)
dagg   <- aggregate(cbind(y, x) ~ id, data = d, FUN = mean)

between <- feols(y ~ x,        data = dagg)
pooled  <- feols(y ~ x,        data = d, cluster = ~id)
within  <- feols(y ~ x | id,   data = d)
mundlak <- feols(y ~ x + xbar, data = d, cluster = ~id)

pull <- function(m, v) c(coef(m)[[v]], se(m)[[v]])
tab <- rbind(`between (collapsed)` = pull(between, "x"),
             `pooled OLS`          = pull(pooled,  "x"),
             `within / FE`         = pull(within,  "x"),
             `Mundlak: x`          = pull(mundlak, "x"),
             `Mundlak: xbar`       = pull(mundlak, "xbar"))
colnames(tab) <- c("estimate", "std. error")
round(tab, 3)
```

Read the table down.  The collapsed regression returns about 1.85, badly
biased, because it is the between estimator and $\bar x_i$ carries $c_i$
with it.  Pooled OLS returns about 1.48 --- closer, since some within
variation is mixed in, but still inconsistent.  The within estimator
returns about 0.99, and only the long format can compute it: once we
collapse there is nothing left to demean.  Mundlak delivers both at
once, exactly as described above --- 0.99 on $x_{it}$, 0.85 on
$\bar x_i$, and $0.99+0.85$ recovers the between estimate.

So the argument for the long format is not precision.  It is that
demeaning, the Hausman comparison, unit-specific slopes, lagged
dependent variables and anything else built on within variation stop
existing once the panel is collapsed.  Dynamic models are the extreme
case, and they get their own chapter,
[Dynamic Panel Data](dynamic-panel.qmd).

## How OLS weights the data

Twice above we said OLS weights units unequally: by $T_i$ when the panel
is unbalanced, and by variance when it mixes within and between
variation.  Neither is a panel-specific quirk.  Both follow from one
rule.

Let $V$ be the regressor of interest and $\tilde V$ be $V$ after
residualising on every other regressor in the model.  By the
Frisch--Waugh--Lovell theorem,

$$
\hat\beta = \frac{\sum_i \sum_t \tilde V_{it}\, y_{it}}
                 {\sum_i \sum_t \tilde V_{it}^{2}} .
$$

Be careful what $\tilde V_{it}$ is.  A single observation has no
variation of its own --- variation belongs to a collection.  What it has
is a **deviation**: how far its regressor sits from what the rest of the
model predicts for it.  With only an intercept that is distance from the
grand mean; with unit effects it is distance from the unit's own mean.
Which mean gets subtracted is the single most consequential detail in
this section, and we return to it.

### Two ways to read the coefficient

The formula supports two readings.  They describe one estimator and
differ in what is being weighted.

$$
\hat\beta = \sum_i \omega_i\, y_i , \quad
\omega_i = \frac{\tilde V_i}{\sum_j \tilde V_j^2}
\qquad\text{versus}\qquad
\hat\beta = \frac{\sum_i w_i\, \beta_i}{\sum_i w_i} , \quad
w_i = \tilde V_i V_i .
$$

| | Weights on outcomes, $\omega_i$ | Weights on effects, $w_i$ |
|---|---|---|
| Weights what | the observed $y_i$ | that observation's own effect $\beta_i$ |
| Sign | signed, sums to zero | non-negative |
| Computable | yes, $y$ is observed | no, the $\beta_i$ are never seen |
| Holds when | always, it is algebra | after writing effects into $y$; the per-unit reading also needs saturation |
| Answers | which rows drive this number, and does any enter with the wrong sign | is this the effect I meant, and whose effect is it mostly |

Default to the second.  Nearly every question that motivates the topic
--- is my coefficient the average effect, whose effect does it report,
why do smaller groups get larger weights --- is a question about the
estimand.  Reach for the first when overlap is in doubt and you need to
see whether observations enter negatively, which squares cannot show.
Both are developed at length in
[Weights in OLS in causal inference](https://xiangao.github.io/blog_book/weights-ols.html).

The square in $w_i$ is derived, not assumed.  Substituting
$y_{it}=\beta_i V_{it}+\ldots$ into the numerator gives
$\sum_i \beta_i \sum_t \tilde V_{it}V_{it}$, and over the whole sample
$\sum \tilde V V=\sum \tilde V^2$ because $\tilde V$ is orthogonal to the
fitted part of $V$.  Whether that holds *unit by unit* --- which is what
licenses reading $\sum_t \tilde V_{it}^2$ as unit $i$'s weight ---
depends on the residualising being done within the unit.  Demeaning
delivers it by direct expansion,
$\sum_t (V_{it}-\bar V_i)V_{it}=\sum_t V_{it}^2 - T\bar V_i^{\,2}
=\sum_t (V_{it}-\bar V_i)^2$, while a coarser control matches only in
total.  Hazlett and Shinkre make this the centre of their treatment; the
Angrist row below assumes the saturated case.

In general $\sum_t \tilde V_{it}^2$ will not factor into anything
tidier, since $\tilde V_{it}$ carries both subscripts even when $V$ does
not.  What changes from model to model is whether it happens to
simplify.

| Model | $\sum_t \tilde V_{it}^2$ simplifies to |
|---|---|
| $z_i$ alone, nothing else varying within the unit | $T_i\,\tilde z_i^2$ |
| a time-varying $x_{it}$, unit effects removed | $(T-1)\,\widehat{\text{Var}}_i(x)$ |
| binary $d$ with saturated controls | $n_x\,\hat d(x)\,(1-\hat d(x))$ |

Add a time-varying $x_{it}$ alongside $z_i$ in the first model and the
simplification is gone: residualising $z_i$ on something that moves
within the unit makes $\tilde z$ move within the unit too.  The rule
survives, the shortcut does not, and the $T_i$ form is the one-regressor
special case rather than the general one.  The last row is Angrist
([1998](https://doi.org/10.2307/2998558)), whose strata weights
$\hat d(x)(1-\hat d(x))$ are largest where treatment is closest to a
coin flip.  Same rule, different regressor.

### When the weighting changes the answer

The weighting always happens.  It changes the answer only when the
effect is heterogeneous *and* the heterogeneity lines up with the
weight.  The claim to check is that a regression does not report *the*
effect but a weighted average of the effects, and that these are
different numbers.

The individual effects are never observed, so we plant them: give every
unit its own slope $\beta_i$, generate $y$ from those slopes, and see
which average of the planted slopes the estimate lands on.  In the next
chunk only one estimator is run; anything else reported is not an
estimate but the planted slopes averaged some way.  Fixed effects is
used because in a panel the natural place for heterogeneity is across
units, and it is still OLS --- OLS with the unit dummies partialled out.

Start with the extreme case.  Half the units barely move and have a true
slope of zero; the other half move a great deal and have a true slope of
two.  The average unit's slope is one.

```{r}
set.seed(1)
half <- N / 2
g <- data.frame(id   = rep(1:N, each = T),
                move = rep(c(rep(0.2, half), rep(3, half)), each = T),
                b    = rep(c(rep(0,   half), rep(2, half)), each = T))
g$x <- rnorm(N * T, 0, g$move)
g$y <- g$b * g$x + rep(rnorm(N), each = T) + rnorm(N * T)

wg <- tapply(resid(feols(x ~ 1 | id, g))^2, g$id, sum)
round(cbind(value = c(
  `average of the two true slopes`      = mean(c(0, 2)),
  `FE estimate`                         = coef(feols(y ~ x | id, g))[["x"]],
  `weight held by the steady half (%)`  = 100 * sum(wg[1:half]) / sum(wg))), 3)
```

The average slope is one and the regression reports 1.98, because the
five hundred steady units hold under half of one percent of the weight.
Their six observations sit almost on top of their own unit mean ---
this regression carries unit effects, so that is the relevant deviation
--- and squaring puts them nowhere in the sum.  Nothing has gone wrong
in the arithmetic.  The units with a slope of zero were never really in
the regression.

### Which mean, and why it matters

That result is about the *within-unit* deviation.  Plain OLS uses
distance from the grand mean instead, and the two coincided above only
because both halves were built around zero.  Separate them and the
algebra says exactly what happens.

Take two equally sized groups with means $\mu_g$, within-group variances
$\sigma_g^2$, slopes $\beta_g$ and group mean outcomes $\bar y_g$.  The
law of total covariance splits $\text{Cov}(x,y)$ and $\text{Var}(x)$
into within and between parts, giving

$$
\text{plim}\;\hat\beta_{OLS}
= \frac{\tfrac12\!\left(\beta_1\sigma_1^2+\beta_2\sigma_2^2\right)
        + \tfrac14\,\Delta\mu\,\Delta\bar y}
       {\tfrac12\!\left(\sigma_1^2+\sigma_2^2\right)
        + \tfrac14\,\Delta\mu^{2}} ,
$$

with $\Delta\mu=\mu_1-\mu_2$ and $\Delta\bar y=\bar y_1-\bar y_2$: a
variance-weighted average of
$\beta_{within}=(\beta_1\sigma_1^2+\beta_2\sigma_2^2)/(\sigma_1^2+\sigma_2^2)$
and $\beta_{between}=\Delta\bar y/\Delta\mu$, weighted by the within
variance $\tfrac12(\sigma_1^2+\sigma_2^2)$ and the between variance
$\tfrac14\Delta\mu^2$.

Everything follows without running anything.  Fixed effects estimates
$\beta_{within}$, here $(0\cdot 0.04 + 2\cdot 9)/9.04 = 1.991$, which
involves no $\mu_g$ --- so parking a group anywhere cannot move it.
Plain OLS also carries $\beta_{between}$, which is zero in this design
because the steady slope and both intercepts are zero, leaving the two
group means at the same height.  As $\Delta\mu$ grows, the between
weight $\tfrac14\Delta\mu^2$ swamps the fixed within weight of $4.52$
and the coefficient is dragged from $1.991$ toward zero like
$9/(4.52+\Delta\mu^2/4)$, the numerator being
$\tfrac12(\beta_1\sigma_1^2+\beta_2\sigma_2^2)=9$.  The simulation is a
check on that, not the source of it.

```{r}
b_within <- (0 * 0.2^2 + 2 * 3^2) / (0.2^2 + 3^2)   # variance-weighted slope
V_within <- 0.5 * (0.2^2 + 3^2)                     # within variance
predicted <- function(shift)                        # beta_between = 0 here
  V_within * b_within / (V_within + 0.25 * shift^2)

park <- function(shift) {
  set.seed(1)
  x <- rnorm(N * T, 0, g$move) + rep(c(rep(shift, half), rep(0, half)), each = T)
  y <- g$b * x + rep(rnorm(N), each = T) + rnorm(N * T)
  d <- data.frame(id = g$id, x = x, y = y)
  c(`FE (theory 1.991)` = coef(feols(y ~ x | id, d))[["x"]],
    `OLS predicted`     = predicted(shift),
    `OLS actual`        = coef(lm(y ~ x, d))[["x"]])
}

round(t(sapply(c(`shift 0` = 0, `shift 10` = 10, `shift 100` = 100), park)), 3)
```

Predicted and actual agree in every row.  Fixed effects is flat at 1.984
against a theoretical 1.991, the gap being sampling noise.  Plain OLS
runs from 1.983 to 0.004 on identical true slopes, partly because the
steady half gains weight as it leaves the grand mean, and partly because
a second channel opens that is not weighting at all: with the halves at
different levels of $x$, part of the fitted line is a comparison
*between* them.  So "the weight is how much $x$ moves" describes the
demeaned case; it is exact once unit effects are removed and false in
general.

### When it does not matter

Weighting only distorts when the effect is related to the weight.  Give
every unit its own slope, let units differ in how much $x$ moves, and
compare two designs: in (A) the slopes are unrelated to that movement,
in (B) the slope is larger where $x$ moves more.  Both average near one.

```{r}
set.seed(7)
sdX <- runif(N, 0.2, 3)          # how much x moves, unit by unit

panel <- function(b) {           # build a panel whose unit slopes are b
  h   <- data.frame(id = rep(1:N, each = T),
                    x  = rnorm(N * T, 0, rep(sdX, each = T)))
  h$y <- rep(b, each = T) * h$x + rep(rnorm(N), each = T) + rnorm(N * T)
  h
}
compare <- function(h, b) {
  w <- tapply(resid(feols(x ~ 1 | id, h))^2, h$id, sum)   # the weights
  c(`FE estimate`            = coef(feols(y ~ x | id, h))[["x"]],
    `true slopes, plain avg` = mean(b),
    `true slopes, weighted`  = sum(w * b) / sum(w))
}

b1 <- rnorm(N, 1, 1)             # (A) slopes unrelated to movement
b2 <- 1 + 0.8 * (sdX - 1.6)      # (B) slopes larger where x moves more
h1 <- panel(b1); h2 <- panel(b2)

round(cbind(`(A) unrelated` = compare(h1, b1),
            `(B) aligned`   = compare(h2, b2)), 4)
```

In (A) all three numbers agree to within sampling noise: the weighting
is there and costs nothing.  In (B) the estimate is 1.53 while the plain
average of the true slopes is 1.01, and the weighted average, 1.54, is
what the estimator was tracking all along.  None of this is a defect ---
a unit whose $x$ barely moves genuinely carries little information about
how $y$ responds, and leaning on the units that do move is efficient.

Słoczyński ([2022](https://doi.org/10.1162/rest_a_00953)) makes the same
point for a binary treatment
([discussed here](https://xiangao.github.io/blog_book/ols-ate.html)):
writing $\rho$ for the treated share, the coefficient is
$(1-\rho)\,\text{ATT} + \rho\,\text{ATU}$, so the *smaller* group
receives the *larger* weight.  In a sample that is 10% treated the
weight on the ATT is $0.9$, so the coefficient mostly reports the effect
on the treated tenth, not the untreated majority.  The ATE is the same
two numbers with the weights reversed, and the two can be far apart: in
a simulation with 8.8% treated, an ATT of 2.39 and an ATU of 0.87, the
ATE is 1.00 while OLS returns 2.26.

With homogeneous effects the weighting is harmless.  With heterogeneous
effects the question is never whether OLS is weighting, only whether we
know what it weighted.
